---
title: "Adoption Decision Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adoption Decision Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide helps decide whether `gamlss.longitudinal` should be the primary
model, a sensitivity model, or not used for a longitudinal analysis. The short
version is:

- use a GEE when the estimand is a marginal mean and the response distribution
  is routine;
- use a GLMM when subject-specific latent heterogeneity is the scientific
  estimand;
- use `gamlss.longitudinal` when the applied answer depends on margins,
  heteroskedasticity, quantiles, tails, asymmetry, or longitudinal dependence.

That framing is deliberately conservative. The package should replace GEE and
GLMM workflows where distributional and dependence modelling changes the answer,
not by pretending every longitudinal question needs a richer model.

## Decision Table

| Applied question | Default starting point | Why |
|---|---|---|
| Average marginal treatment effect only | GEE | The GEE estimand matches the question and is fast. |
| Subject-specific random-effect interpretation | GLMM | The random effect is part of the scientific model. |
| Mean plus calibrated intervals or quantiles | `gamlss.longitudinal` | Model `mu`, `sigma`, and other margin parameters directly. |
| Heteroskedastic longitudinal outcome | `gamlss.longitudinal` | Put variance drivers in `sigma.formula` and check interval calibration. |
| Positive, skewed, count, or tail-sensitive outcome | `gamlss.longitudinal` | Choose a GAMLSS family and inspect quantile/tail diagnostics. |
| Time-varying dependence | `gamlss.longitudinal` | Put time or covariates in `theta.formula`. |
| Sparse data with weakly identified distributional parameters | GEE/GLMM plus sensitivity analysis | Richer margins may be unstable. |
| Regulatory or clinical report requiring a familiar primary model | GEE/GLMM plus `gamlss.longitudinal` sensitivity | Keep the primary estimand familiar and report distributional sensitivity. |

## Translation Table

| GEE/GLMM concept | GAMLSS-copula concept |
|---|---|
| Mean formula | `mu.formula` |
| Working correlation | Copula family plus `theta.formula` |
| Random intercept as dependence device | Copula dependence parameter |
| Overdispersion or residual scale | `sigma.formula` |
| Shape, zero inflation, tail behaviour | GAMLSS family parameters such as `nu` or `tau` |
| Robust mean-focused diagnostics | `check_model()` marginal, PIT, tail, and dependence decisions |
| Mean prediction | `predict(type = "response")` |
| Prediction interval or percentile | `predict(type = "quantile")` |
| Subject trajectory simulation | `simulate(type = "copula")` |

## Minimum Evidence For Replacement

A replacement claim needs more than a successful fit. Before making
`gamlss.longitudinal` the primary model, check:

1. The estimand needs distributional or dependence modelling.
2. `check_model()` does not flag unresolved marginal, tail, or dependence
   problems.
3. The familiar GEE or GLMM does not answer the same applied question more
   simply.
4. Scenario-level benchmarks or sensitivity analyses support the replacement
   for the relevant estimand.
5. The report states whether inference is likelihood/Hessian based,
   delta-method, bootstrap/simulation based, or approximate.

Use `vignette("inference-uncertainty", package = "gamlss.longitudinal")` as the
reporting checklist for that last item.

The benchmark framework is meant to support this evidence:

```{r benchmark-report}
bench <- run_adoption_benchmarks(
  scenarios = c("gaussian_heteroskedastic", "time_varying_dependence"),
  reps = 100,
  methods = c("rs_separate", "gee", "glmm", "gam"),
  max_elapsed_sec = 60,
  write_results = TRUE
)

write_benchmark_report(bench)
```

Read the scenario-level sections of the report first. They are restricted to the
declared primary metrics for each scenario, which keeps the claim tied to the
problem being tested.

## When To Use It As The Primary Model

Use `gamlss.longitudinal` as the primary model when one or more of these are
true:

- clinical or scientific decisions depend on quantiles, tails, or calibrated
  probabilities;
- scale, skewness, zero inflation, or tail behaviour changes with covariates;
- residual dependence remains after a mean-only GEE or random-intercept GLMM;
- time-varying or covariate-varying dependence is scientifically relevant;
- simulation, prediction, or subject trajectory generation is part of the
  deliverable;
- scenario-level benchmarks show wins or meaningful ties on the relevant
  estimand.

## When To Use It As Sensitivity Analysis

Use it as a sensitivity model when the GEE or GLMM remains the clearest primary
analysis, but distributional assumptions still matter. Good sensitivity uses
include:

- checking whether a marginal mean conclusion changes under a richer response
  family;
- verifying whether tails or intervals are underfit by a standard model;
- exploring whether dependence is more structured than a random intercept or
  exchangeable working correlation;
- reporting subject-trajectory simulations alongside a familiar primary model.

In this role, the model earns trust by explaining what the simpler model misses,
not by replacing it automatically.

## When Not To Use It

Do not make `gamlss.longitudinal` the primary model when:

- the scientific question is only a marginal mean and a GEE fits cleanly;
- a random effect is the scientific target rather than a nuisance dependence
  device;
- the sample is too small to identify the chosen distributional parameters;
- the time grid or missingness mechanism requires a model outside the current
  adjacent-time copula workflow;
- diagnostics show unresolved marginal, tail, or dependence misfit;
- benchmark or sensitivity results show no practical gain over the familiar
  model;
- stakeholders need a simple estimand and the richer model would obscure the
  answer.

In those cases, a GEE or GLMM may be the better primary analysis. The
GAMLSS-copula model can still be valuable as a transparent sensitivity check.

## Suggested Reporting Language

Use specific language:

- "We used a GAMLSS-copula model as the primary model because the estimand
  included calibrated upper-tail probabilities."
- "The GEE was retained as the primary mean model; the GAMLSS-copula analysis was
  used as a distributional sensitivity analysis."
- "The GAMLSS-copula model improved interval calibration in the scenario-level
  benchmark, but the GEE remained faster and comparable for the marginal mean."
- "We did not use the GAMLSS-copula model as primary because diagnostics showed
  unresolved tail misfit."

Avoid broad language:

- "GAMLSS-copula models replace GEEs."
- "The benchmark proves this package always beats GLMMs."
- "A more flexible distribution is automatically better."

The adoption case is strongest when it is narrow, diagnostic-driven, and tied to
the estimand.
