---
title: "Benchmarking Adoption Claims"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking Adoption Claims}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide is for building evidence that `gamlss.longitudinal` should replace
a GEE or GLMM in a specific class of longitudinal problems. It is not meant to
produce one universal leaderboard. Each benchmark scenario names the applied
claim, the data-generating setting, the comparator methods, and the metrics
that decide whether the claim is supported.

For reporting language around model checks and uncertainty, see
`vignette("diagnostics-decisions", package = "gamlss.longitudinal")` and
`vignette("inference-uncertainty", package = "gamlss.longitudinal")`.

## Scenario Plan

Start by inspecting the built-in adoption scenarios:

```{r scenarios}
scenarios <- adoption_benchmark_scenarios()
scenarios[, c("scenario", "family", "copula", "design", "missingness", "claim")]
```

The initial scenarios are deliberately narrow:

- Gaussian outcomes with a true scale-varying margin, where calibration matters.
- Positive skewed outcomes where mean-only models can miss quantiles and tails.
- Count outcomes where dispersion and upper-tail behaviour matter.
- Gaussian outcomes where adjacent-time dependence changes over follow-up.
- Missing-visit panels where the workflow should remain stable.

These are the first places where a GAMLSS-copula model should be able to make a
credible adoption case. If it only ties on those scenarios, the package should
be described as a sensitivity model rather than a replacement.

## Smoke Run

Use a tiny run while developing benchmark code:

```{r smoke}
smoke <- run_adoption_benchmarks(
  scenarios = "gaussian_heteroskedastic",
  reps = 2,
  methods = c("rs_separate", "gee", "glmm", "gam"),
  max_elapsed_sec = 20,
  write_results = FALSE
)

smoke
smoke$summary$summary
```

This is a wiring check only. It confirms that the scenario can be simulated,
each comparator can be attempted, and the result table can be summarized.

## Full Opt-In Run

Run the full benchmark outside CRAN checks and ordinary unit tests:

```{r full-run}
bench <- run_adoption_benchmarks(
  reps = 100,
  methods = c("rs_separate", "gee", "glmm", "gam"),
  max_elapsed_sec = 60,
  write_results = TRUE
)

bench$summary$summary
write_benchmark_report(bench)
```

The saved result rows are intentionally verbose. They include convergence,
runtime, likelihood summaries where meaningful, truth-aware mean error,
model-implied 90th percentile error, upper-tail error at the true 90th
percentile, residual-calibrated interval diagnostics for Gaussian baselines,
PIT calibration, and tail errors where available.

`write_benchmark_report()` creates a Markdown report with both aggregate and
scenario-level sections. Use the scenario-level sections for adoption claims:
they are filtered to each scenario's declared primary metrics, so a count or
positive-outcome scenario is not accidentally interpreted through a Gaussian
interval metric that was not part of its claim.

## Minimum Standard For A Public Claim

Treat small benchmark runs as wiring checks. A public package-site or manuscript
claim should be based on a locked scenario plan, a recorded comparator
availability table, and enough repetitions that the scenario-level conclusions
are stable.

Before writing that `gamlss.longitudinal` replaces a GEE or GLMM for a problem
class, save:

- the scenario table from `adoption_benchmark_scenarios()`;
- the package availability table from `benchmark_comparator_status()`;
- the raw result CSV and RDS from `run_adoption_benchmarks()`;
- the Markdown report from `write_benchmark_report()`;
- the scenario-level primary metrics used for the claim;
- the cases where a simpler GEE or GLMM remains faster, clearer, or equally
  accurate for the target estimand.

This keeps the adoption claim narrow enough to trust.

## Reading The Summary

`summarise_benchmark_results()` labels each method within each scenario and
replicate as a win, tie, loss, or missing result for each metric. The scoring
rules depend on the estimand:

- Mean errors and runtime are lower-is-better.
- 90th percentile error is lower-is-better.
- Time-varying dependence error is lower-is-better.
- Bias and tail errors target zero.
- 95% interval coverage targets 0.95.
- PIT Kolmogorov-Smirnov p-values are higher-is-better.

That means a method can win one estimand and lose another. This is a feature,
not a nuisance: it prevents a mean-only model from being declared the winner
when it predicts the mean well but misses tail behaviour.

## Reporting Claims

Use benchmark language that matches the evidence:

```{r report-table}
bench$summary$summary[
  bench$summary$summary$metric %in% c(
    "benchmark_mean_rmse",
    "benchmark_q90_mae",
    "benchmark_upper_tail_error_90",
    "benchmark_interval_coverage_95",
    "benchmark_tail_error_upper_05",
    "elapsed_sec"
  ),
]
```

For scenario-specific claims, inspect the Markdown report or the case-level
table:

```{r scenario-report}
report <- format_benchmark_report(bench)
grep("Scenario-Level Headline Results", report, value = TRUE)

bench$summary$case_results[
  bench$summary$case_results$benchmark_scenario == "time_varying_dependence" &
    bench$summary$case_results$metric == "benchmark_theta_time_abs_error",
]
```

Good benchmark claims are specific:

- "The GAMLSS-copula workflow wins or ties on interval calibration and tail
  behaviour in the Gaussian heteroskedastic scenario."
- "The GLMM is faster, but tail diagnostics show undercoverage in the positive
  skewed scenario."
- "The benchmark is inconclusive for count outcomes because a comparator failed
  often enough to reduce the effective comparison set."

Avoid broad claims such as "this package beats GLMMs". Broad claims are hard to
defend and easy to misread. The goal is to show where a richer longitudinal
distributional model changes the applied answer.

## Current Extended-Run Lesson

A local 20-replicate run across the five built-in scenarios is useful as a
directional check, not as final evidence. In that run all 400 fits succeeded.
The scenario-level report suggested:

- `rs_separate` was the only method scoring the time-varying dependence
  estimand, because the standard baselines do not model changing copula
  dependence.
- `rs_separate` led the heteroskedastic Gaussian scenario on the 90th-percentile
  metric and was competitive on mean RMSE.
- `rs_separate` led the missing-visits scenario on the declared interval and
  mean metrics.
- GEE remained the fastest method and was strongest in the positive and count
  scenarios under these simple data-generating mechanisms.

That is exactly the kind of result the benchmark framework should make visible:
the package has a clearer adoption case when distributional calibration or
dependence is the estimand, while simple marginal-mean use cases may still
prefer a GEE.
