---
title: "Diagnostics As Decisions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics As Decisions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Diagnostics should decide what happens next. In `gamlss.longitudinal`, the
first diagnostic pass is `check_model()`: it combines numerical summaries,
plain-language basic checks, and failed-check warnings. Use `plot(fit)` and the
explicit diagnostic helpers for visual review.

The goal is not to create a longer residual-plot ritual. The goal is to answer:

- is the marginal distribution credible enough for the estimand?
- are tails and quantiles calibrated well enough to report?
- does dependence remain after the copula?
- is inference stable enough to cite?
- did the basic automated checks pass, or do any areas need review?

## Run The Diagnostic Check

```{r check}
check <- check_model(
  fit,
  include_vcov = TRUE
)

check
check$basic_checks
check$basic_checks_result
check$warnings
plot(fit)
```

`check$basic_checks` contains the compact five-row status table printed at the
top of the output. `check$checks` contains the detailed backing table, including
rows where no automatic check was triggered. `check$warnings` contains failed
checks only; review rows remain visible in the check tables but do not emit an R
warning.

```{r check-tables}
check$basic_checks
check$checks
check$warnings
```

## Marginal Calibration

The marginal distribution is the foundation of a GAMLSS-copula model. If the
margin is wrong, dependence and tail conclusions can also be wrong.

Use:

```{r marginal-calibration}
check$pit
pithist(fit)
qqrplot(fit)
wormplot(fit)
```

Decision language:

- PIT uniformity concerns mean the chosen family or covariate formulas are not
  explaining the response distribution well enough.
- A pattern in the QQ or worm plot points to systematic underfit, not a random
  nuisance.
- Revise `margin_dist`, `mu.formula`, `sigma.formula`, or shape-parameter
  formulas before making distributional claims.

Possible next actions:

- try a skewed, heavy-tailed, zero-inflated, or count-specific family;
- add known scale drivers to `sigma.formula`;
- simplify unstable smooths before adding more parameters;
- keep a GEE or GLMM as primary if the distributional model does not improve
  the applied answer.

## Tail Calibration

Tail diagnostics matter when reporting quantiles, threshold probabilities,
extreme outcomes, or safety events.

```{r tail-calibration}
check$tail
rootogram(fit)
predict(fit, type = "quantile", probs = c(0.05, 0.5, 0.95))
```

Decision language:

- upper-tail underfit means high outcomes occur more often than the fitted
  margin expects;
- lower-tail underfit means low outcomes occur more often than expected;
- tail warnings should be resolved before using the model as primary for tail
  probabilities.

Possible next actions:

- use a heavier-tailed or asymmetric family;
- add distributional covariates to scale or shape parameters;
- report tail estimates as sensitivity-only if diagnostics remain unresolved.

## Residual Dependence

The copula is meant to absorb longitudinal association left after the margin.
Residual dependence warnings mean the chosen copula family or `theta.formula`
may be too simple.

```{r dependence}
check$residual_dependence
check$copula
copula_time_summary(fit)
plot_copula_diagnostics(fit, data = dat)
```

Decision language:

- residual dependence after the copula means subject trajectories are still
  more associated than the model expects;
- time-varying association should usually be tested with `theta.formula = ~ time`
  or another scientifically motivated dependence formula;
- a random-intercept GLMM or exchangeable GEE may remain a useful sensitivity
  model, but it does not answer the same dependence question as a time-varying
  copula.

Possible next actions:

- compare copula families with `select_copula()`;
- add time or baseline predictors to `theta.formula`;
- inspect scenario-level benchmarks for `benchmark_theta_time_abs_error` when
  dependence recovery is the target.

## Inference Warnings

Inference warnings are about whether standard errors, intervals, and tests are
stable enough to report.

```{r inference}
summary(fit, include_vcov = TRUE)
confint(fit)
check$warnings[check$warnings$area == "inference", ]
```

Decision language:

- analytical Hessian inference is the preferred fast likelihood-based route
  when available;
- numerical Hessian inference should be described as approximate;
- convergence, identifiability, or boundary concerns should be reported and
  investigated, not hidden.

For detailed reporting language, use
`vignette("inference-uncertainty", package = "gamlss.longitudinal")`.

## Basic Check Result

Use the diagnostic result as a first-pass screen, not as a replacement for
visual and subject-matter diagnostics.

`check$basic_checks_result` gives a compact first pass:

- `passed` means all five basic automated checks returned `PASS`;
- `review` means no check failed, but at least one check returned `REVIEW`;
- `failed` means at least one check returned `FAIL` and `check_model()` emitted
  an R warning.

| Diagnostic state | Suggested role |
|---|---|
| Basic checks passed; estimand needs distribution/tails/dependence | Primary model candidate |
| Mean is stable but tails or dependence remain unresolved | Sensitivity model |
| Marginal calibration is poor | Revise before interpreting |
| Dependence remains after the copula | Revise copula or `theta.formula` |
| Inference is fragile in a small sample | Sensitivity model plus simulation |
| GEE/GLMM answers the estimand more simply | Keep GEE/GLMM primary |

## Reporting Template

A concise diagnostics paragraph can look like this:

> We reviewed marginal calibration, tail calibration, residual dependence, and
> inference diagnostics using `check_model()`. The model showed no unresolved
> marginal or tail warnings, and normal-score residual dependence was small
> after the fitted copula. Coefficient intervals used the reported
> variance-covariance route; numerical-Hessian intervals, if used, were treated
> as approximate.

If diagnostics do flag problems, state them plainly:

> The GAMLSS-copula model was retained as a sensitivity analysis because
> `check_model()` flagged unresolved upper-tail misfit. The GEE remained the
> primary analysis for the marginal mean estimand.

This is the trust-building habit: diagnostics should either support the richer
model or explain why a simpler model remains primary.
