---
title: "Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

We provide a range of diagnostics for assessing the validity of `gamlss.longitudinal` fits, generally grouped by whether they are marginal diagnostics (which are similar to many other marginal modelling package) or diagnostics for the joint fit which includes dependence.

We provide an example review of key diagnostics for an example dataset as a part of the detailed workflow here: [Detailed worked example](native-simulation-workflow.html). The goal of this page is as a reference to outline the available methods and provide brief details on how they operate.

In general we suggest the following checks for any model:

- `check_model(fit)` provides descriptive numerical summaries of the marginal and dependence fit and is a good starting point for review
- `summary(fit)` naturally is an important starting point for reviewing fit likelihood, coefficients and their significance to them model
- `plot_terms(fit)` provides a visual assessment of the fitted model coefficients and their confidence intervals
- `plot(fit)` provides a faceted plot of visual marginal fit diagnostics
- `plot_copula_diagnostics(fit)` provides a plot of visual joint distribution and dependence fit diagnostics


## Basic checks

We provide a brief diagnostic summary using `check_model()`. It is deliberately descriptive and should be used alongside graphical diagnostics and subject-matter review. The package does not convert PIT or tail summaries into pass/fail decisions.

| Area | Quantity reported | Default interpretation | Status |
|---|---|---|---|
| Convergence | Fitted-object convergence contract | Required before inference | `converged` or `not_converged` |
| Marginal fit | PIT KS summary | Descriptive; inspect PIT/QQ/worm/rootogram plots | `descriptive` |
| Tail fit | Lower/upper PIT tail ratios | Descriptive; interpret in context | `descriptive` |
| Copula fit | Rosenblatt residual correlation by requested lag | Descriptive by default | `descriptive` |
| Variance calculation | Variance-covariance method provenance | Report the inference contract | `recorded`, `review`, or `not_computed` |

For discrete margins, `check_model()` uses randomized PIT values with a recorded seed and preserves the caller's random-number state. Dependence lags without usable pairs are retained with an explicit `scope_status`. A user may supply `dependence_cor_cutoff`, but any resulting `flagged` status is a user-defined review flag, not a formal test or a package verdict.

```{r check}
check <- check_model(fit)

check
check$basic_checks
check$checks
check$diagnostic_summary
check$flags
```

## Marginal distribution checks

The marginal fit affects both the margin and the copula, as the copula relies on the results from the margin fit to strip out marginal effects from dependence. The most straightforward way to review marginal fit is with `plot(fit)`.

`plot(fit)` creates diagnostic plots for:

- Probability integral transform (PIT) histogram (checking if the fitted model results in uniform fitted CDF)
- Quantile-quantile (QQ) plot of the fitted versus empirical distribution (points should follow the diagonal line)
- Wormplot, which can be viewed as a 'higher resolution' version of a QQ-plot which includes reference ranges and should have points within the reference ranges, ideally without persistent trends away from the centre line 
- Rootogram, showing the difference in the binned distribution for the observed versus fitted distribution (these should centre on zero without significant trending deviation)

The default `plot()` function plots the diagnostics over the whole distribution, but there is the option to facet the plots by time to assess fit across the longitudinal timepoints to review any issues over repeated observations.

```{r marginal-diagnostics-bytime}
rootogram(fit, by_time = TRUE)
pithist(fit, by_time = TRUE)
qqrplot(fit, by_time = TRUE)
wormplot(fit, by_time = TRUE)
```

We also provide the option to facet each of the diagnostics by a covariate using `by="covariate_name"`. Note that when including a variable to facet by, it's required to provide the original dataset as we don't expose all covariates as part of the fitted object.

```{r marginal-diagnostics-bycovariate}
pithist(fit, by="treatment", data=dataset)
qqrplot(fit, by="treatment", data=dataset)
wormplot(fit, by="treatment", data=dataset)
rootogram(fit, by="treatment", data=dataset)
```

If any of the above diagnostics are of concern then marginal fit needs to be adjusted. Review options for `margin_dist` or coefficient fits for available parameters `mu.formula`, `sigma.formula`, `nu.formula` and/or `tau.formula`.

## Dependence structure diagnostics

The copula structure specified by `copula_dist` should capture the shape of the dependence after the marginal shape is removed, with `theta.formula` and `zeta.formula` allowing the shape to vary with time or other covariates.

In the same way as a margin, we expect the residual trends remaining in our dependence structure after imposing our dependence fit to be essentially random,

`plot_copula_diagnostics(fit)` provides a detailed set of diagnostics for the dependence structure, including:

- Fitted copula overlayed on a histogram of the dependence (fitted contours should follow density of the histogram)
- Quantiles (deciles) of fitted Kendall's tau (correlation) for the fitted versus observed copula (observed and fitted should broadly be in line)
- Rosenblatt z-scores by timepoint and overall Rosenblatt normal QQ which should be normally distributed (these take into account the correlation structure for timepoints beyond the first)
- Rosenblatt lag plot showing Rosenblatt z-scores for each time point against the lagged timepoint which should should zero correlation, as well as residual dependence by lag which should show close to zero residual correlation
- Sorted empirical versus fitted copula probabilities scatterplot which should closely track the centre line
- Tail co-occurrence and exceedance, essentially the fitted versus observed probability of observations landing in the upper and lower tails of the dependence structure

```{r dependence}
plot_copula_diagnostics(fit, data = dat)
```

Any concerns in these diagnostics indicate lack of fit for the dependence, or potentially, the margin fit affecting the dependence fit. If marginal checks are reasonable, then adjustments to copula should include assessing alternative copula shapes, i.e. `copula_dist`, or adjusting parameter covariates for `theta.formula` or `zeta.formula`.
