---
title: "Inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article gives an overview of the outputs that can be generated from a `gamlss.longitudinal` fit for inference of various types. 

The quick reference is:

- (ideally) use `check_model()` first to make sure there aren't any major errors flagged in the fitted model that may effect inference
- use `summary()` and `confint()` for fixed-coefficient estimates and conditional standard errors and intervals
- use `wald_test()` for coefficient hypotheses
- use `likelihood_compare()` for likelihood ratio comparisons for models (ideally nested)
- use `bootstrap_inference()` for bootstrap standard errors and confidence intervals for coefficients
- use `predict(..., se.fit = TRUE, interval = "confidence")` for conditional,
  fixed-`mu` confidence limits, and use fitted quantiles for plug-in descriptions
  of response variation; these are different from future-observation prediction intervals
- use `marginal_effects(..., se.fit = TRUE)` for averaged response against covariates
- use `simulate()` for simulating replicates from the fitted model

## Variance-covariance and standard errors

`summary(fit)` reports fixed-coefficient estimates with standard errors. The variance covariance can be requested directly using the standard call, `vcov(fit)`. By default these methods use the Hessian for all fixed margin and copula coefficients, calculated using primarily analytical methods (i.e. second derivatives of the known margin and copula distributions) with small exceptions where analytical methods are not available and numerical methods are required. This covariance is conditional on the fitted smooth coefficients, penalties, and smoothing parameters. Smooth covariance is calculated separately from penalized working weights as an approximate conditional quantity. Neither result includes fixed-smooth cross-covariance or smoothing-parameter uncertainty.

The complete machine-readable inventory is available from `inference_contracts()`. Inference-producing objects carry an `inference_contract` attribute (or component for covariance and summary objects) recording the target coefficient names, conditioning, omissions, approximation, and fit-level validity status.

```{r summary}
s <- summary(fit, include_vcov = TRUE)
s$coefficients
s$fit$vcov_method
```

```{r vcov}
vc <- vcov(fit, method = "analytical")
vc$method
vc$method_requested
```

If the analytical Hessian calculation fails, or the user selects `vcov(fit, method = "numderiv")` or `summary(fit, vcov_method="numderiv")`, the package will use a fully numerically estimated Hessian and records the method used. The `check_model(fit)` function will also provide a warning and flag status of the model as 'review' rather than passed or failed if the Hessian failed to calculate analytically.

`vcov(fit, method = "numderiv")` requests numerical-Hessian inference directly if preferred, or of interest to compare to the analytical method. Our standard package tests verify the two methods are within a small tolerance of one another at build time for a selection of fits, so generally the results should closely agree.

```{r numderiv}
vc_num <- vcov(fit, method = "numderiv")
vc_num$method
```

Since we are using the inverted Hessian directly, these intervals are conditional large-sample approximations. A comparison with bootstrap intervals can assess sensitivity to the local quadratic approximation, but does not add model-selection uncertainty or turn the result into unconditional all-parameter inference.

## Confidence intervals for coefficients

Use `confint()` for coefficient-scale intervals.

```{r confint}
confint(fit)
confint(fit, parm = c("mu.treatment", "sigma.treatment"))
```

If the fitted object reports convergence, Hessian, identifiability, or fallback
warnings, these should ideally be rectified or carefully reported alongside model fit.

## Wald tests

The `wald_test()` can be used to test coefficient estimates against an alternative hypothesis, and defaults to an individual z-test for single parameters.

```{r wald-individual}
wald_test(
  fit,
  terms = c("mu.treatment", "sigma.treatment"),
  method = "analytical"
)
```

We also provide the option to test a group of coefficients together against a null hypothesis using `joint = TRUE` which results in a chi-square test.

```{r wald-joint}
wald_test(
  fit,
  terms = c("mu.treatment", "sigma.treatment"),
  joint = TRUE,
  method = "analytical"
)
```

Note that we are assuming asymptotic normality for the estimator of the parameters and their covariance (for multiple parameters) which may not necessarily be true for these joint distributions.

## Likelihood comparisons

Compare alternative `gamlss.longitudinal` fits with `likelihood_compare(fit,fit2,fit3...)`. 

```{r likelihood-compare}
likelihood_compare(
  reduced = fit_reduced,
  full = fit
)
```

For the likelihood ratio to follow the theoretical Chi-squared distribution, the models should be nested, i.e. the parameters of one model are a subset of the compared model. If the models are not nested the result should be considered as approximate.

The comparison reports likelihood and likelihood ratio test results alongside other helpful details such as the change in degrees of freedom and other likelihood criteria that take into account effective degrees of freedom: AIC and BIC.

## Parametric bootstrap

We provide a built-in helper for bootstrap simulations from fitted `gamlss.longitudinal` objects which can give an idea of sensitivity of the Hessian-based intervals. The parametric method simulates from the fitted model and refits the same specification; the cluster method resamples whole subjects. Both target only the selected fixed coefficients. Refit errors, explicit nonconvergence, and missing or nonfinite target coefficients count as failed replicates. Results report overall and per-term successful counts, and model-selection uncertainty is not included.

```{r bootstrap-inference}
boot <- bootstrap_inference(
  fit,
  R = 200,
  terms = c("mu.treatment", "sigma.treatment"),
  seed = 1,
  fit_args = list(max_outer_iter = 20, max_inner_iter = 20)
)

boot
boot$summary
```

These fits are computationally expensive due to the refitting of the `gamlss.longitudinal` model so use with care.

## Prediction of response or coefficients

The `predict` function provides row-level fitted marginal quantities. `predict(type = "mean")` returns fitted response means and `predict(type = "mu")` returns the fitted distribution parameter. With `se.fit = TRUE` or `interval = "confidence"`, the current delta method uses only fixed `mu` coefficients, conditional on all other parameters and fitted smooths. It omits non-`mu`, smooth, fixed-smooth, and smoothing-parameter uncertainty. These limits are confidence limits for the fitted quantity, not prediction intervals for a future response.

```{r predict-response}
p <- predict(
  fit,
  newdata = dat,
  type = "mean",
  se.fit = TRUE,
  interval = "confidence",
  level = 0.95
)
head(p)
```

`predict()` returns a marginal fit and does not condition on a subject's other responses or preserve a simulated trajectory. The fitted dependence model affects the estimated coefficients and their fixed-coefficient covariance, but the returned confidence limits do not represent trajectory dependence.

If the quantity of interest is the dependence structure then simulation from the fitted object via `simulate(fit)` may provide the required functionality.

## Prediction of quantiles or tail probabilities

More flexible distributional models such as GAMLSS provide fitted marginal quantiles and probabilities beyond the mean. `predict(type = "quantile")`,
`predict(type = "cdf")`, `predict(type = "density")`, and `predict(type = "probability")` can give a view of these quantities for analysis and reporting on a row level.

These distributional predictions are plug-in point summaries: they hold every fitted parameter fixed and do not include parameter, smooth, or model-selection uncertainty. A pair of fitted quantiles describes model-implied response variation; it is not a confidence interval for an estimated quantile.

```{r predict-distribution}
predict(fit, newdata = dat, type = "quantile", probs = c(0.1, 0.5, 0.9))
predict(fit, newdata = dat, type = "cdf", q = 10)
predict(fit, newdata = dat, type = "density", y = 10)
predict(fit, newdata = dat, type = "probability", q = 10, direction = "above")
```

## Averaged predictions (marginal effects)

`marginal_effects()` reports averaged fitted values across supplied counterfactual rows. With `se.fit = TRUE`, its uncertainty is an exploratory aggregation of rowwise fixed-`mu` delta-method standard errors. It omits cross-row covariance as well as non-`mu`, smooth, fixed-smooth, and smoothing-parameter uncertainty, so it should not be presented as unconditional all-parameter inference.

```{r marginal-effects}
eff <- marginal_effects(
  fit,
  newdata = dat,
  variable = "treatment",
  parameter = "mu",
  se.fit = TRUE
)
eff
```

## Dependence predictions

As noted previously, `predict()` provides marginal fitted quantities; its conditional fixed-`mu` confidence limits do not preserve dependence for new trajectories.

If the subject of interest is behaviour of dependence under new datasets or forecast trajectories into later timepoints, then a simulation-based approach may be more appropriate.

We provide tooling to simulate directly from a fitted `gamlss.longitudinal` fit with `simulate(fit)` or `simulate(fit,type="copula")`.

```{r simulate}
sim <- simulate(
  fit,
  nsim = 100,
  seed = 1
)
```

More details on simulation is available in the reference guide for [Simulator usage](simulator-usage.html).
