---
title: "Inference and Uncertainty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inference and Uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide explains what can be reported from a `gamlss.longitudinal` fit and
how to describe the uncertainty honestly. It is meant to support applied reports
where users need language as familiar as GEE robust standard errors or GLMM
Wald intervals, while still being clear about the richer likelihood and
copula-based model.

The short version is:

- use `summary()` and `confint()` for coefficient-scale inference;
- use `wald_test()` for individual or joint coefficient hypotheses;
- use `likelihood_compare()` for nested likelihood-ratio model comparisons;
- use `bootstrap_inference()` for opt-in parametric bootstrap coefficient
  checks;
- use `predict(..., se.fit = TRUE, interval = "confidence")` for fast
  response-mean intervals;
- use `marginal_effects(..., se.fit = TRUE)` for counterfactual contrasts;
- use `simulate()` for trajectory, tail, and dependence uncertainty checks;
- use `check_model(..., include_vcov = TRUE)` to make inference warnings visible
  before reporting results.

## Fit With Inference In Mind

The high-level workflow keeps variance-covariance computation optional because
large longitudinal likelihoods can make Hessian work expensive. For final
reporting, either compute it at fit time or request it when summarising.

```{r fit}
fit <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = gamlss.dist::GA(mu.link = "log", sigma.link = "log"),
  copula_dist = "C",
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + s(age_scaled, bs = "ps"),
  sigma.formula = ~ treatment + time,
  theta.formula = ~ time,
  compute_vcov = TRUE,
  vcov_method = "analytical"
)
```

For exploratory model building, it is reasonable to leave `compute_vcov = FALSE`
and request inference only for shortlisted models:

```{r late-vcov}
summary(fit, include_vcov = TRUE)
vc <- vcov(fit, method = "analytical")
confint(fit, method = "analytical")
wald_test(fit, terms = c("mu.treatment", "sigma.treatment"))
likelihood_compare(fit_reduced, fit)
bootstrap_inference(fit, R = 200, terms = "mu.treatment")
```

## What The Methods Mean

`summary(fit, include_vcov = TRUE)` reports coefficient estimates with standard
errors when a variance-covariance matrix is available. The default route is
likelihood-based Hessian inference using the analytical Hessian path where
possible.

```{r summary}
s <- summary(fit, include_vcov = TRUE)
s$coefficients
s$fit$vcov_method
```

`vcov(fit, method = "analytical")` requests the analytical Hessian path. If that
path fails and fallback is allowed, the package falls back to the numerical
Hessian path and records the method used.

```{r vcov}
vc <- vcov(fit, method = "analytical")
vc$method
vc$method_requested
```

`vcov(fit, method = "numderiv")` requests numerical-Hessian inference directly.
This is useful when the analytical path is not available for the fitted family,
parameterisation, or local numerical state.

```{r numderiv}
vc_num <- vcov(fit, method = "numderiv")
vc_num$method
```

These intervals are large-sample likelihood approximations. They are not GEE
sandwich standard errors, and they are not subject-level bootstrap intervals
unless a separate bootstrap or simulation analysis is performed.

## Coefficient Intervals

Use `confint()` for coefficient-scale intervals.

```{r confint}
confint(fit)
confint(fit, parm = c("mu.treatment", "sigma.treatment"))
```

Suggested reporting language:

> Coefficient intervals were computed from the fitted likelihood using the
> package variance-covariance method reported by `summary()`. Numerical Hessian
> intervals are treated as approximate.

If the fitted object reports convergence, Hessian, identifiability, or fallback
warnings, include them in the results section rather than hiding them in a
supplement.

## Wald Tests

Use `wald_test()` for coefficient-scale hypothesis tests. With `terms`, the
default is an individual Wald z-test for each selected coefficient.

```{r wald-individual}
wald_test(
  fit,
  terms = c("mu.treatment", "sigma.treatment"),
  method = "analytical"
)
```

Set `joint = TRUE` to test a group of coefficients together with a Wald
chi-square test.

```{r wald-joint}
wald_test(
  fit,
  terms = c("mu.treatment", "sigma.treatment"),
  joint = TRUE,
  method = "analytical"
)
```

For custom linear hypotheses, supply a contrast matrix. Columns can be named
with the relevant coefficient names.

```{r wald-contrast}
L <- matrix(c(1, -1), nrow = 1)
colnames(L) <- c("mu.treatmentB", "mu.treatmentC")

wald_test(fit, L = L, method = "analytical")
```

Suggested reporting language:

> Wald tests used the fitted model variance-covariance matrix reported by
> `vcov()`. Tests using the numerical Hessian route were treated as approximate.

## Likelihood Comparisons

Use `likelihood_compare()` for sequential likelihood-ratio summaries across
nested or approximately nested `gamlss.longitudinal` models.

```{r likelihood-compare}
likelihood_compare(
  reduced = fit_reduced,
  full = fit
)
```

The helper reports joint log-likelihood, effective degrees of freedom, AIC, BIC,
the likelihood-ratio increment, and a chi-square reference p-value for each
comparison against the previous row. It warns when models use different
observation counts.

Suggested reporting language:

> Likelihood-ratio comparisons used the fitted joint likelihood and chi-square
> reference tests for nested model contrasts. We treated comparisons as
> approximate when models differed in smoothing structure or were not strictly
> nested.

## Parametric Bootstrap

Use `bootstrap_inference()` when Hessian-based intervals need a model-based
sensitivity check. The helper simulates responses from the fitted model, refits
the same model to each simulated response, and summarizes percentile intervals
for selected fixed coefficients.

```{r bootstrap-inference}
boot <- bootstrap_inference(
  fit,
  R = 200,
  terms = c("mu.treatment", "sigma.treatment"),
  seed = 1,
  fit_args = list(max_outer_iter = 20, max_inner_iter = 20)
)

boot
boot$summary
```

This is intentionally opt-in because refitting longitudinal likelihoods can be
expensive. Use enough replicates for final reporting, record failed refits, and
describe the result as a parametric bootstrap under the fitted margin and
copula model.

Suggested reporting language:

> Parametric bootstrap intervals were computed by simulating responses from the
> fitted GAMLSS-copula model and refitting the same model to each simulated
> dataset. Failed refits were recorded and excluded from percentile summaries.

## Response Mean Intervals

For applied interpretation, response-scale estimates are usually clearer than
raw coefficients. Use `predict(type = "response")` for fitted means and request
standard errors when the uncertainty should be shown.

```{r predict-response}
p <- predict(
  fit,
  newdata = dat,
  type = "response",
  se.fit = TRUE,
  interval = "confidence",
  level = 0.95
)
head(p)
```

These are delta-method confidence intervals for the fitted response mean. They
are fast and useful for tables, but they should not be described as prediction
intervals for new longitudinal trajectories.

Suggested reporting language:

> Response-mean confidence intervals were computed on the fitted model scale
> using the delta method and transformed back to the response scale.

## Quantiles And Tail Probabilities

Distributional longitudinal models are often used because the answer depends on
quantiles or probabilities, not just the mean. Use `predict(type = "quantile")`,
`predict(type = "cdf")`, `predict(type = "density")`, and
`predict(type = "probability")` to report those estimands directly.

```{r predict-distribution}
predict(fit, newdata = dat, type = "quantile", probs = c(0.1, 0.5, 0.9))
predict(fit, newdata = dat, type = "cdf", q = 10)
predict(fit, newdata = dat, type = "density", y = 10)
predict(fit, newdata = dat, type = "probability", q = 10, direction = "above")
```

For final tail-sensitive claims, pair these estimates with `check_model()` tail
diagnostics and, when possible, simulation-based sensitivity checks.

```{r check-tail}
check <- check_model(fit, include_vcov = TRUE)
check$tail
check$warnings
```

Suggested reporting language:

> Upper-tail probabilities were estimated from the fitted GAMLSS margin and
> interpreted only after PIT and tail calibration diagnostics were reviewed.

## Marginal Effects

Use `marginal_effects()` for counterfactual contrasts that are easier to explain
than coefficient changes, especially with nonlinear links, non-Gaussian
margins, or distributional parameters.

```{r marginal-effects}
eff <- marginal_effects(
  fit,
  newdata = dat,
  variable = "treatment",
  parameter = "mu",
  se.fit = TRUE
)
eff
```

The uncertainty is variance-covariance based and inherits the same Hessian or
fallback caveats as `summary()`, `vcov()`, and `confint()`.

## Trajectory And Dependence Uncertainty

When the scientific target is a subject trajectory, clustered tail behaviour, or
dependence-sensitive event probability, use simulation from the fitted model.

```{r simulate}
sim <- simulate(
  fit,
  nsim = 100,
  seed = 1,
  type = "copula"
)
```

Simulation does not automatically solve parameter uncertainty. It is best
viewed as fitted-model uncertainty unless paired with a bootstrap or repeated
refitting workflow. It is still the right first tool for checking whether the
model can reproduce longitudinal patterns that coefficient intervals cannot
show.

Suggested reporting language:

> Simulated trajectories were drawn from the fitted margin and copula model to
> assess dependence and tail behaviour implied by the fitted parameters.

## Diagnostic Gatekeeping

Before treating intervals as reportable, run `check_model()` with inference
metadata included.

```{r check}
check <- check_model(fit, include_vcov = TRUE)
check$checks
check$warnings
```

Do not promote a model as the primary analysis if diagnostics still flag
unresolved marginal misfit, residual dependence, or tail calibration problems.
In that case, report the model as exploratory or sensitivity analysis, revise
the margin or copula, or retain the simpler GEE/GLMM as primary.

## Small Samples And Fragile Hessians

Use extra caution when:

- few subjects are available;
- the time grid is short or highly unbalanced;
- a shape or tail parameter is weakly identified;
- the fitted copula dependence is near a boundary;
- analytical Hessian inference falls back to numerical Hessian inference;
- `check_model()` reports convergence or scoring concerns.

In these settings, prefer sensitivity analysis, simulation, and transparent
benchmarking over a single table of Wald-style intervals.

## Reporting Checklist

For final applied reports, include:

1. the estimand: mean, quantile, probability, trajectory, or dependence;
2. the margin family and copula family;
3. whether dependence varies through `theta.formula`;
4. the variance-covariance method used for coefficient or mean intervals;
5. whether intervals are likelihood/Hessian, numerical-Hessian, delta-method, or
   simulation summaries;
6. the main `check_model()` basic check result;
7. any GEE, GLMM, or GAM sensitivity result used as a benchmark.

A compact report sentence might be:

> We used a Gamma GAMLSS-copula model with time-varying copula dependence because
> the estimand included upper-tail probabilities. Coefficient intervals used the
> analytical Hessian variance-covariance path, response-mean intervals used the
> delta method, and diagnostics did not flag unresolved tail or dependence
> misfit.

If that sentence cannot be completed honestly, the model is not ready to be the
primary analysis.
