---
title: "Missing Visits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing Visits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide is for longitudinal panels where subjects miss scheduled follow-up
visits. The goal is to make the missingness explicit before fitting the model,
then use diagnostics and benchmark sensitivity checks to decide whether the
GAMLSS-copula workflow is stable enough to be the primary analysis.

## Start With The Visit Pattern

Before fitting a model, summarize the observed time grid:

```{r visit-pattern}
visit_table <- with(dat, table(subject, time))
visit_counts <- rowSums(visit_table > 0)
table(visit_counts)

missing_by_time <- aggregate(
  response ~ time,
  data = transform(dat, response = !is.na(response)),
  FUN = function(x) mean(!x)
)
missing_by_time
```

Use this step to separate common cases:

- Sporadic missing outcomes on an otherwise regular grid.
- Dropped visit rows where a scheduled time is absent for some subjects.
- Informative dropout, where missingness depends on prior outcomes or risk.

The first two are reasonable starting points for the current workflow. The
third needs sensitivity analysis and may need an explicit missingness model
outside this package.

## Fit The Longitudinal Model

Keep the first model simple. Use formulas that are stable under the observed
visit pattern, then add distributional or dependence complexity only when
diagnostics support it.

```{r fit}
fit <- fit_longitudinal(
  dataset = dat,
  margin_dist = gamlss.dist::NO(),
  copula_dist = "N",
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + age_scaled,
  sigma.formula = ~ treatment + time,
  theta.formula = ~ 1
)
```

If visits are irregular or the time grid is sparse, start with
`theta.formula = ~ 1`. Move to a time-varying dependence formula only when
there are enough adjacent observations to estimate it.

## Diagnostics

```{r check}
check <- check_model(fit)
check$checks
check$residual_dependence
check$tail
```

Focus on three warnings:

- Remaining residual dependence can mean the missing visits changed the
  effective dependence pattern.
- Marginal calibration warnings can mean the response distribution is being
  distorted by selective missingness.
- Tail warnings can mean high-risk subjects are more likely to miss visits or
  drop out.

Use `simulate()` to inspect whether fitted trajectories look plausible under
the observed visit pattern:

```{r simulate}
sim <- simulate(fit, nsim = 100, type = "copula")
```

## Prediction With Missing Visits

Predictions for observed rows are straightforward:

```{r predict-observed}
predict(fit, type = "response", se.fit = TRUE, interval = "confidence")
predict(fit, type = "quantile", probs = c(0.1, 0.5, 0.9))
```

For a complete target grid, build `newdata` deliberately. Do not silently treat
unobserved visits as observed outcomes.

```{r predict-grid}
target_grid <- expand.grid(
  subject = unique(dat$subject),
  time = sort(unique(dat$time)),
  KEEP.OUT.ATTRS = FALSE
)

target_grid <- merge(
  target_grid,
  unique(dat[c("subject", "treatment", "age_scaled")]),
  by = "subject",
  all.x = TRUE
)

predict(fit, newdata = target_grid, type = "response", se.fit = TRUE)
```

## Sensitivity Benchmark

Use the built-in missing-visits adoption scenario as a first wiring check:

```{r benchmark}
missing_bench <- run_adoption_benchmarks(
  scenarios = "missing_visits",
  reps = 50,
  methods = c("rs_separate", "gee", "glmm", "gam"),
  max_elapsed_sec = 60,
  write_results = TRUE
)

missing_bench$summary$summary
```

Read this as a stability check, not proof that missingness is ignorable. If the
model wins or ties under missing visits but fails under complete data, the
problem is probably not missingness. If it is stable under complete data but
fails under missing visits, report the missingness sensitivity directly.

## When To Be Cautious

Be cautious when missingness depends on previous responses, when entire tails of
the outcome distribution are absent after dropout, or when time-varying
dependence is estimated from very few adjacent pairs. In those settings,
`gamlss.longitudinal` can still be a useful outcome model, but the primary
analysis needs a missingness strategy that matches the scientific design.
