---
title: "Native Simulation, Copula Screening, and GAMLSS Longitudinal Fitting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Native Simulation, Copula Screening, and GAMLSS Longitudinal Fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  error = FALSE
)

```

This vignette shows an end-to-end workflow for simulated longitudinal data with
GAMLSS margins and native copula dependence. The example is simulated so that we
can compare the fitted model with the known truth, but the same workflow is
intended to mirror a real analysis:

1. simulate or load longitudinal data,
2. inspect the marginal response and pairwise dependence,
3. choose a marginal distribution,
4. screen candidate copula families,
5. fit the longitudinal model,
6. inspect model, term, and copula diagnostics.

```{r packages}
library(gamlss.longitudinal)
library(gamlss.dist)
library(ggplot2)
```

## Simulate Data

We simulate a balanced dataset with 300 subjects and 5 time points. The response
is Gamma distributed. Each subject has:

- a three-level treatment group: `control`, `low`, or `high`;
- a subject-level age generated uniformly between 25 and 70;
- a repeated observation time scaled to `[0, 1]`.

The marginal mean has treatment, time, and nonlinear age effects. The marginal
scale has treatment, time, and a simple U-shaped age effect. The dependence
structure is Clayton, with its parameter varying smoothly with age and linearly
with time.

```{r simulate-data}
set.seed(20260521)

n_subject <- 300
time_points <- seq(0, 1, length.out = 5)

treatment_effect <- c(control = 0, low = 0.20, high = 0.45)

truth_mu_age_smooth <- function(x) {
  sim_smooth_bump(x, amplitude = 0.45, location = 0.62, width = 0.18) +
    sim_smooth_sin(x, amplitude = 0.12)
}

truth_sigma_age_smooth <- function(x) {
  sim_smooth_u(x, amplitude = 1.00)
}

truth_theta_age_smooth <- function(x) {
  sim_smooth_bump(x, amplitude = 0.55, location = 0.45, width = 0.22)
}

dat <- simulate_longitudinal_dataset(
  n = n_subject,
  times = time_points,
  margin_dist = GA(mu.link = "log", sigma.link = "log"),
  copula_dist = "C",
  covariates = function(base) {
    simulate_longitudinal_covariates(
      base,
      subject = list(
        treatment = function(subject_data) {
          factor(
            sample(c("control", "low", "high"), nrow(subject_data), replace = TRUE),
            levels = c("control", "low", "high")
          )
        },
        age = function(subject_data) stats::runif(nrow(subject_data), 25, 70)
      ),
      observation = list(
        time_scaled = function(long_data) sim_rescale01(long_data$time)
      )
    )
  },
  margin_params = list(
    mu = function(d) {
      age_scaled <- sim_rescale01(d$age)
      eta_mu <- 1.00 +
        sim_factor_effect(d$treatment, treatment_effect) +
        0.20 * d$time_scaled +
        truth_mu_age_smooth(age_scaled)
      exp(eta_mu)
    },
    sigma = function(d) {
      age_scaled <- sim_rescale01(d$age)
      eta_sigma <- -0.75 +
        0.12 * sim_factor_effect(d$treatment, c(control = 0, low = 1, high = 1)) +
        0.10 * d$time_scaled +
        truth_sigma_age_smooth(age_scaled)
      exp(eta_sigma)
    }
  ),
  copula_params = list(
    theta = function(e) {
      age_scaled <- sim_rescale01(e$age)
      time_left_scaled <- sim_rescale01(e$time_left)
      eta_theta <- 0.05 +
        0.20 * time_left_scaled +
        truth_theta_age_smooth(age_scaled)
      exp(eta_theta)
    }
  ),
  seed = 20260521
)

dat$age_scaled <- sim_rescale01(dat$age)
head(dat)
```

## Inspect The Response

`plotDist()` gives a compact view of the marginal response distribution at each
time point and the pairwise dependence between time points. The off-diagonal
panels use pseudo-observations here, so the dependence pattern is shown on a
copula scale.

```{r plot-dist, fig.width = 8, fig.height = 8}
plotDist(dat, GA(mu.link = "log", sigma.link = "log"), offdiag_scale = "pseudo")
```

## Choose A Marginal Distribution

For positive continuous responses, `gamlss::fitDist()` is a useful marginal
screen. It does not account for longitudinal dependence, so it should be treated
as a guide rather than as the final model choice.

```{r marginal-screen-fit, include = FALSE}
quiet_fitDist <- function(...) {
  out_con <- file(tempfile(), open = "wt")
  msg_con <- file(tempfile(), open = "wt")
  sink(out_con)
  sink(msg_con, type = "message")
  on.exit({
    sink(type = "message")
    sink()
    close(msg_con)
    close(out_con)
  }, add = TRUE)
  suppressWarnings(suppressMessages(gamlss::fitDist(...)))
}

margin_screen <- quiet_fitDist(
    dat$response,
    type = "realplus",
    try.gamlss = FALSE,
    trace = FALSE
)
```

```{r marginal-screen}
margin_screen_fits <- data.frame(
  family = names(margin_screen$fits),
  AIC = as.numeric(margin_screen$fits),
  row.names = NULL
)

head(margin_screen_fits, 10)

best_margin <- margin_screen_fits[1, , drop = FALSE]
best_margin
```

The screen is marginal, and this example continues with the Gamma family because
the fitted longitudinal model will handle covariates, smooth effects, and
within-subject dependence.

```{r marginal-histdist, fig.width = 7, fig.height = 4.5}
gamlss::histDist(
  dat$response,
  family = GA,
  nbins = 30,
  main = "Gamma Fit To The Marginal Response",
  xlab = "Response"
)
```

## Choose A Copula Family

The native `select_copula()` helper screens candidate copula families using
copula log-likelihoods on pseudo-observation pairs. In real data, those
pseudo-observations would usually come from a preliminary marginal fit. Because
this example is simulated, we can use the stored truth uniforms to demonstrate
the family-screening step directly.

```{r copula-screen}
copula_family_screen <- select_copula(
  data = dat,
  u_var = "u",
  families = c("N", "C", "F", "G", "J", "t"),
  t_df_grid = c(3, 4, 6, 10, 20)
)

copula_family_screen
```

The next plot overlays the selected copula density on a binned view of adjacent
pairs after transforming the pseudo-observations to standard-normal scale.

```{r copula-screen-overlay, fig.width = 6.5, fig.height = 5}
make_adjacent_pairs <- function(data, subject_var = "subject", time_var = "time", u_var = "u") {
  data <- data[order(data[[subject_var]], data[[time_var]]), , drop = FALSE]
  pair_list <- lapply(split(data, data[[subject_var]]), function(x) {
    x <- x[order(x[[time_var]]), , drop = FALSE]
    data.frame(
      u1 = x[[u_var]][-nrow(x)],
      u2 = x[[u_var]][-1]
    )
  })
  do.call(rbind, pair_list)
}

plot_copula_screen_overlay <- function(pair_data, selected_fit, grid_n = 80, bins = 28) {
  pair_data$z1 <- stats::qnorm(pmin(pmax(pair_data$u1, 0.001), 0.999))
  pair_data$z2 <- stats::qnorm(pmin(pmax(pair_data$u2, 0.001), 0.999))

  z_grid <- seq(-2.75, 2.75, length.out = grid_n)
  density_grid <- expand.grid(z1 = z_grid, z2 = z_grid)
  density_grid$u1 <- stats::pnorm(density_grid$z1)
  density_grid$u2 <- stats::pnorm(density_grid$z2)
  rho <- sin(pi * selected_fit$tau[1] / 2)
  density_grid$density <- exp(
    stats::dnorm(density_grid$z1, log = TRUE) +
      stats::dnorm(
        density_grid$z2,
        mean = rho * density_grid$z1,
        sd = sqrt(1 - rho^2),
        log = TRUE
      )
  )

  ggplot(pair_data, aes(x = z1, y = z2)) +
    geom_bin2d(aes(fill = after_stat(density)), bins = bins) +
    geom_contour(
      data = density_grid,
      aes(x = z1, y = z2, z = density),
      inherit.aes = FALSE,
      color = "white",
      linewidth = 0.75,
      bins = 8
    ) +
    scale_fill_viridis_c(name = "Empirical density") +
    labs(
      title = paste("Selected copula:", selected_fit$family[1], "(Gaussian tau-matched reference overlay)"),
      x = expression(Phi^-1 * (U[t])),
      y = expression(Phi^-1 * (U[t + 1]))
    ) +
    theme_minimal()
}

pair_data <- make_adjacent_pairs(dat)
plot_copula_screen_overlay(pair_data, copula_family_screen[1, , drop = FALSE])
```

## Fit The Longitudinal Model

The final model uses smooth age effects in the marginal mean, marginal scale,
and copula dependence parameter. The example keeps `sigma` slightly less
flexible than `mu` and `theta`, because scale smooths are harder to recover from
finite samples.

```{r fit-model}
fit <- gamlss.longitudinal(
  dataset = dat,
  margin_dist = GA(mu.link = "log", sigma.link = "log"),
  copula_dist = "C",
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time_scaled + s(age_scaled, bs = "ps", k = 8),
  sigma.formula = ~ treatment + time_scaled + s(age_scaled, bs = "ps", k = 5),
  theta.formula = ~ time_scaled + s(age_scaled, bs = "ps", k = 8)
)

summary(fit)
```

The standard model diagnostic dashboard combines PIT, QQ, worm, rootogram, and
forecast-style checks.

```{r model-diagnostics, fig.width = 8, fig.height = 9}
plot(fit, data = dat)
```

## Term Diagnostics

Because this is simulated, we first plot the true term contributions used to
generate the data. The fitted term dashboard is shown immediately afterwards.

```{r true-terms, fig.width = 8, fig.height = 5}
term_theme <- function(p) {
  p +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 11),
      axis.title = element_text(size = 9),
      plot.caption = element_text(size = 8)
    )
}

truth_smooth_plot <- function(x, contribution, title, xlab) {
  plot_df <- data.frame(x = x, contribution = contribution)
  term_theme(
    ggplot(plot_df, aes(x = x, y = contribution)) +
      geom_line(color = "black", linewidth = 1.2) +
      labs(title = title, x = xlab, y = paste0("smooth(", xlab, ")"), caption = "truth")
  )
}

truth_continuous_plot <- function(x, contribution, title, xlab, ylab) {
  plot_df <- data.frame(x = x, contribution = contribution)
  term_theme(
    ggplot(plot_df, aes(x = x, y = contribution)) +
      geom_line(color = "black", linewidth = 1.2) +
      labs(title = title, x = xlab, y = ylab, caption = "truth")
  )
}

truth_factor_plot <- function(effects, title, xlab, ylab) {
  plot_df <- data.frame(
    x = seq_along(effects),
    level = factor(names(effects), levels = names(effects)),
    contribution = as.numeric(effects)
  )
  term_theme(
    ggplot(plot_df, aes(x = x, y = contribution)) +
      geom_hline(yintercept = 0, color = "grey70", linetype = 3) +
      geom_point(color = "black", size = 1.2) +
      geom_errorbar(aes(ymin = contribution, ymax = contribution), color = "red", width = 0.15) +
      scale_x_continuous(breaks = plot_df$x, labels = plot_df$level) +
      labs(title = title, x = xlab, y = ylab, caption = "truth")
  )
}

print_term_dashboard <- function(plotlist, ncol = 4) {
  nrow <- ceiling(length(plotlist) / ncol)
  grid::grid.newpage()
  grid::pushViewport(grid::viewport(layout = grid::grid.layout(nrow, ncol)))
  on.exit(grid::popViewport(), add = TRUE)
  for (i in seq_along(plotlist)) {
    row_id <- ((i - 1) %/% ncol) + 1
    col_id <- ((i - 1) %% ncol) + 1
    print(
      plotlist[[i]],
      vp = grid::viewport(layout.pos.row = row_id, layout.pos.col = col_id)
    )
  }
}

time_grid <- seq(0, 1, length.out = 200)
age_grid <- seq(0, 1, length.out = 200)

truth_term_plots <- list(
  truth_smooth_plot(age_grid, truth_mu_age_smooth(age_grid), 'mu: s(age_scaled, bs = "ps", k = 8)', "age_scaled"),
  truth_smooth_plot(age_grid, truth_sigma_age_smooth(age_grid), 'sigma: s(age_scaled, bs = "ps", k = 5)', "age_scaled"),
  truth_smooth_plot(age_grid, truth_theta_age_smooth(age_grid), 'theta: s(age_scaled, bs = "ps", k = 8)', "age_scaled"),
  truth_factor_plot(treatment_effect, "mu: treatment", "treatment", "fixed contribution: mu.treatment"),
  truth_continuous_plot(time_grid, 0.20 * time_grid, "mu: time_scaled", "time_scaled", "fixed contribution: mu.time_scaled"),
  truth_factor_plot(c(control = 0, low = 0.12, high = 0.12), "sigma: treatment", "treatment", "fixed contribution: sigma.treatment"),
  truth_continuous_plot(time_grid, 0.10 * time_grid, "sigma: time_scaled", "time_scaled", "fixed contribution: sigma.time_scaled"),
  truth_continuous_plot(time_grid, 0.20 * time_grid, "theta: time_scaled", "time_scaled", "fixed contribution: theta.time_scaled")
)

print_term_dashboard(truth_term_plots, ncol = 4)
```

```{r fitted-terms, fig.width = 8, fig.height = 5}
plot_terms(fit, data = dat)
```

## Copula Diagnostics

Finally, inspect the fitted copula directly. The default dashboard includes
normal-scale pair diagnostics, fitted copula overlays, tail summaries, and
lagged residual checks.

```{r copula-diagnostics, fig.width = 8, fig.height = 8}
plot_copula_diagnostics(fit, data = dat)
```

## Next Steps

In a real analysis, the same workflow can be repeated with alternative marginal
formulas, different copula families, and more targeted diagnostics. For
simulation studies, the stored `true_*` columns make it straightforward to
compare fitted effects against the data-generating truth.
