---
title: "Detailed worked example for a longitudinal gamlss model fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detailed worked example for a longitudinal gamlss model fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  out.width = "92%",
  dpi = 180,
  fig.retina = 2,
  warning = FALSE,
  message = FALSE,
  error = FALSE
)

```

```{css, echo = FALSE}
.figure {
  text-align: center;
}

.figure img {
  display: block;
  margin-left: auto;
  margin-right: auto;
  max-width: 100%;
}

details {
  margin: 1rem 0;
}

details summary {
  cursor: pointer;
  font-weight: 600;
}
```

This vignette shows an end-to-end workflow for fitting a simulated longitudinal dataset with a longitudinal gamlss model with gamlss margins and copula dependence using **gamlss.longitudinal**.

The example is simulated so that we can compare the fitted model with the known truth, but the same workflow would be intended for a real dataset (excluding simulation of the dataset, of course).

The workflow is:

0. simulate a longitudinal dataset using `simulate_longitudinal_dataset()`,
1. inspect the marginal response, covariates, missingness, and pairwise dependence,
2. select a marginal distribution and copula family,
3. fit the longitudinal model,
4. review inference and model diagnostics,
5. answer applied questions with prediction, marginal effects, simulation, and
   benchmark comparisons.

Because this is a regression workflow, the code below separates exploratory
screening from modelling decisions. In a real analysis, the main estimand,
scientifically important covariates, factor reference levels, transformations,
and sensitivity checks should be chosen before treating coefficient estimates
as confirmatory evidence.

```{r packages}
library(gamlss.longitudinal)
library(ggplot2)
```

## Simulate Data

We simulate a balanced dataset with 120 subjects and 5 time points. The response
is generalized gamma distributed, with 10% response missingness added at random
conditional on observed time, treatment, and age. Each subject has:

- a three-level treatment group: `control`, `low`, or `high`;
- a subject-level age generated uniformly between 25 and 70;
- a repeated observation time scaled to `[0, 1]`.

The marginal mean has treatment, time, and nonlinear age effects. The marginal
scale has deliberately visible treatment, time, and age effects so the fitted
term diagnostics have something meaningful to recover. A generalized gamma
shape term pushes the marginal distribution away from a plain Gamma screen. The
dependence structure is Clayton, with its parameter increasing strongly over
time so adjacent visits become more correlated later in follow-up.

<details>
<summary>Simulation code</summary>

```{r simulate-data}
set.seed(20260521)

n_subject <- 120
time_points <- seq(0, 1, length.out = 5)

treatment_effect <- c(control = 0, low = 0.30, high = 0.65)

truth_mu_age_smooth <- function(x) {
  sim_smooth_bump(x, amplitude = 0.60, location = 0.62, width = 0.18) +
    sim_smooth_sin(x, amplitude = 0.16)
}

truth_sigma_age_smooth <- function(x) {
  sim_smooth_u(x, amplitude = 0.45)
}

truth_nu_age_smooth <- function(x) {
  sim_smooth_sigmoid(x, amplitude = 0.35, location = 0.45, steepness = 9)
}

truth_theta_age_smooth <- function(x) {
  sim_smooth_bump(x, amplitude = 0.25, location = 0.45, width = 0.22)
}

dat <- simulate_longitudinal_dataset(
  n = n_subject,
  times = time_points,
  margin_dist = GG(mu.link = "log", sigma.link = "log", nu.link = "identity"),
  copula_dist = "C",
  covariates = function(base) {
    simulate_longitudinal_covariates(
      base,
      subject = list(
        treatment = function(subject_data) {
          factor(
            sample(c("control", "low", "high"), nrow(subject_data), replace = TRUE),
            levels = c("control", "low", "high")
          )
        },
        age = function(subject_data) stats::runif(nrow(subject_data), 25, 70)
      ),
      observation = list(
        time_scaled = function(long_data) sim_rescale01(long_data$time)
      )
    )
  },
  margin_params = list(
    mu = function(d) {
      age_scaled <- sim_rescale01(d$age)
      eta_mu <- 1.00 +
        sim_factor_effect(d$treatment, treatment_effect) +
        0.35 * d$time_scaled +
        truth_mu_age_smooth(age_scaled)
      exp(eta_mu)
    },
    sigma = function(d) {
      age_scaled <- sim_rescale01(d$age)
      eta_sigma <- -1.05 +
        0.25 * sim_factor_effect(d$treatment, c(control = 0, low = 1, high = 1)) +
        1.25 * d$time_scaled +
        truth_sigma_age_smooth(age_scaled)
      exp(eta_sigma)
    },
    nu = function(d) {
      age_scaled <- sim_rescale01(d$age)
      1.15 + 0.55 * d$time_scaled + truth_nu_age_smooth(age_scaled)
    }
  ),
  copula_params = list(
    theta = function(e) {
      age_scaled <- sim_rescale01(e$age)
      time_left_scaled <- sim_rescale01(e$time_left)
      eta_theta <- -0.70 +
        1.80 * time_left_scaled +
        truth_theta_age_smooth(age_scaled)
      exp(eta_theta)
    }
  ),
  seed = 20260521
)

dat$age_scaled <- sim_rescale01(dat$age)

missing_weight <- stats::plogis(
  -0.70 +
    1.15 * dat$time_scaled +
    0.45 * (dat$treatment == "high") -
    0.30 * dat$age_scaled
)
missing_rows <- sample(
  seq_len(nrow(dat)),
  size = round(0.10 * nrow(dat)),
  replace = FALSE,
  prob = missing_weight
)
dat$response[missing_rows] <- NA_real_
dat$response_missing <- is.na(dat$response)

head(dat)
```

</details>

For real data, this is also the point to check the analysis-ready data frame:
one row per subject-time combination, stable subject identifiers, correctly
ordered time values, intended factor reference levels, and missing response
patterns. Continuous variables used in smooths or interactions are often easier
to interpret and fit when centered or scaled before modelling.

```{r data-shape}
# Check data columns / subject/time counts
str(dat)
table(dat$treatment)
table(dat$time)
mean(dat$response_missing)

aggregate(true_sigma ~ time, dat, mean)
aggregate(true_nu ~ time, dat, mean)
aggregate(true_theta ~ time, dat, mean, na.rm = TRUE)
```

`check_missingness()` screens whether response missingness is associated with
observed features. A significant predictor association is compatible with MAR
conditional on those features; it does not prove MNAR, and a nonsignificant
screen cannot rule MNAR out.

```{r missingness-check}
missingness_check <- check_missingness(
  dat,
  response_var = "response",
  time_var = "time",
  subject_var = "subject",
  predictors = c("time_scaled", "treatment", "age_scaled")
)

missingness_check
```

## Inspect The Response

`plot_dist()` gives a compact view of the marginal response distribution at each
time point and the pairwise dependence between time points. The off-diagonal
panels use pseudo-observations here, so the dependence pattern is shown on a
copula scale. In this simulation, the diagonal panels should become more
right-skewed over time, and later adjacent-visit panels should show stronger
dependence.

```{r plot-dist, fig.width = 8.5, fig.height = 8.5, out.width = "88%", fig.alt = "Matrix plot of marginal response distributions and pairwise dependence by time."}
plot_dist(
  dat,
  margin_dist = GG(mu.link = "log", sigma.link = "log", nu.link = "identity"),
  subject_var = "subject",
  time_var = "time",
  response_var = "response"
)
```

## Choose A Marginal Distribution

For positive continuous responses, `select_margin()` gives a ranked marginal
screen and marks whether each family is supported by
`gamlss_longitudinal()`. Because this simulated longitudinal dataset has strong
time-varying scale and shape, we score the retained families with
time-specific intercepts for each distribution parameter. This better matches
the final model structure than a single intercept-only fit to the pooled
response distribution. The screen still does not account for longitudinal
dependence or non-time covariates, so it is a shortlist rather than the final
model.

```{r marginal-screen}
margin_candidate_families <- c(
  "GG", "GA", "GIG", "WEI", "EXP"
)

margin_selection <- select_margin(
  dat,
  response_var = "response",
  time_var = "time",
  time_intercepts = TRUE,
  type = "realplus",
  families = margin_candidate_families,
  try.gamlss = FALSE,
  trace = FALSE
)

margin_selection

margin_choice <- best_fit(margin_selection)
margin_choice[c("family_name", "rank", "AIC", "delta_AIC", "supported_by_longitudinal")]

screened_margin_family <- margin_choice$family_name
screened_margin_dist <- best_fit_family(margin_selection)
margin_dist <- if (identical(screened_margin_family, "GG")) {
  GG(mu.link = "log", sigma.link = "log", nu.link = "identity")
} else {
  screened_margin_dist
}
gamma_margin_dist <- GA(mu.link = "log", sigma.link = "log")
```


```{r marginal-fit-overlay, fig.width = 9.5, fig.height = 7.5, out.width = "95%", fig.alt = "Histograms of observed responses comparing generalized gamma and gamma marginal fits with time-intercept overlays."}
margin_overlay_plots <- list(
  plot_margin_fit(
    dat,
    margin_dist = margin_dist,
    response_var = "response",
    bins = 30,
    time_intercepts = TRUE,
    time_var = "time",
    plot = FALSE
  )$plot,
  plot_margin_fit(
    dat,
    margin_dist = gamma_margin_dist,
    response_var = "response",
    bins = 30,
    time_intercepts = TRUE,
    time_var = "time",
    plot = FALSE
  )$plot
)

ggpubr::ggarrange(plotlist = margin_overlay_plots, ncol = 2, nrow = 1)
```

The pooled overlays above are now mixtures across the fitted time-specific
margins. The faceted view below uses the same time-intercept marginal fit, but
shows each time point separately so changes in scale and shape are easier to
inspect.

```{r marginal-fit-overlay_long, fig.width = 9.5, fig.height = 7.5, out.width = "95%", fig.alt = "Histograms of fitted GG with time-varying intercepts for each parameter."}
plot_margin_fit(
  dat, 
  margin_dist = margin_dist, 
  time_intercepts = TRUE, 
  by_time=TRUE, 
  time_var="time", 
  response_var="response"
)
```

## Choose A Copula Family

The native `select_copula()` helper screens candidate copula families using
copula log-likelihoods on adjacent-time pseudo-observation pairs. In the
standard workflow, pseudo-observations are created automatically from the
observed response and selected marginal distribution.

```{r copula-screen}
copula_selection <- select_copula(
  data = dat,
  response_var = "response",
  margin_dist = margin_dist,
  subject_var = "subject",
  time_var = "time",
  families = c("N", "C", "F", "G", "J", "t"),
  t_df_grid = c(3, 4, 6, 10, 20)
)

copula_selection

copula_choice <- best_fit(copula_selection)
copula_choice[c("family", "rank", "AIC", "delta_AIC", "tau")]

screened_copula_dist <- best_fit_family(copula_selection)
copula_dist <- screened_copula_dist
```

The separate marginal and copula screens are intentionally lightweight. As an
additional screen, `select_joint_distribution()` ranks margin-copula
combinations by fitting simple joint longitudinal models. This is slower, but
it confirms whether the separate screens point in the same direction.

```{r joint-screen}
top_margin_families <- head(
  margin_selection$family[margin_selection$supported_by_longitudinal],
  10
)
joint_screen_margins <- unique(c(screened_margin_family, "GA"))

joint_selection <- select_joint_distribution(
  data = dat,
  response_var = "response",
  subject_var = "subject",
  time_var = "time",
  margin_families = joint_screen_margins,
  copula_families = c("N", "C", "F", "G", "J", "t"),
  fit_args = list(
    max_outer_iter = 2,
    max_inner_iter = 2,
    outer_stop_crit = 1e-2,
    inner_stop_crit = 1e-2,
    use_backtracking = FALSE
  ),
  trace = FALSE
)

joint_selection

joint_choice <- best_fit(joint_selection)
joint_choice[c("margin_family_name", "copula_family", "rank", "AIC", "delta_AIC")]

joint_gamma_comparison <- aggregate(
  AIC ~ margin_family,
  data = as.data.frame(joint_selection),
  FUN = min
)
joint_gamma_comparison$delta_AIC <- joint_gamma_comparison$AIC -
  min(joint_gamma_comparison$AIC, na.rm = TRUE)
joint_gamma_comparison[order(joint_gamma_comparison$AIC), ]

data.frame(
  screen = c("separate", "joint"),
  margin = c(screened_margin_family, joint_choice$margin_family_name),
  copula = c(copula_dist, joint_choice$copula_family)
)
```

## Fit The Longitudinal Model

The final model uses smooth age effects in the marginal mean, marginal scale,
generalized-gamma shape, and copula dependence parameter. The example makes the
time effects on `mu` and `sigma`, and the treatment effect on `sigma`,
intentionally strong enough that the fitted term diagnostics can recover them
with useful uncertainty bands.

Before fitting, decide what the regression model is meant to estimate. Here the
primary questions are how treatment and time affect the marginal mean, whether
the scale changes across treatment, time, and age, and whether within-subject
dependence changes with time and age. The factor levels were fixed at
simulation time so `control` is the reference group, and `age_scaled` was put on
`[0, 1]` so smooths and copula terms are easier to fit and compare. The spline
basis dimensions (`k`) are deliberately modest: large enough to represent the
known smooths, but not so large that the scale and dependence submodels become
hard to interpret.

```{r fit-model, results='hide'}
fit <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = margin_dist,
  copula_dist = copula_dist,
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time_scaled + s(age_scaled, bs = "ps", k = 8),
  sigma.formula = ~ treatment + time_scaled + s(age_scaled, bs = "ps", k = 5),
  nu.formula = ~ time_scaled + s(age_scaled, bs = "ps", k = 5),
  theta.formula = ~ time_scaled + s(age_scaled, bs = "ps", k = 8),
  max_outer_iter = 8,
  max_inner_iter = 8,
  outer_stop_crit = 1e-3,
  inner_stop_crit = 1e-3,
  verbose = 0
)
```

```{r fit-summary}
summary(fit)
```

The summary gives coefficient estimates, uncertainty, and model fit statistics.
The marginal mean (`mu`), marginal scale (`sigma`), generalized-gamma shape
(`nu`), and copula dependence (`theta`) coefficients answer different
questions, so they should not be read as a single "effect" table. Mean effects
describe the response location, scale effects describe heterogeneity or
dispersion, shape effects describe remaining distributional asymmetry, and
dependence effects describe within-subject association after the marginal model
has been accounted for.

The fitted marginal overlay now uses the row-specific distributional parameters
from the longitudinal model, so it is a better check than the intercept-only
screening plot for whether the fitted marginal distribution tracks the observed
response scale and spread. We show the fitted overlay on both response and
log-response scales.

```{r fitted-margin-overlay, fig.width = 9.5, fig.height = 4.75, out.width = "95%", fig.alt = "Histograms of observed responses with fitted longitudinal generalized gamma marginal density overlays on response and log-response scales."}
fitted_margin_overlay_plots <- list(
  plot_margin_fit(
    fit = fit,
    data = dat,
    bins = 30,
    plot = FALSE
  )$plot,
  plot_margin_fit(
    fit = fit,
    data = dat,
    bins = 30,
    response_scale = "log",
    plot = FALSE
  )$plot
)

ggpubr::ggarrange(plotlist = fitted_margin_overlay_plots, ncol = 2, nrow = 1)
```

```{r inference}
coef_names <- names(coef(fit))
inference_terms <- coef_names[
  grepl("^mu\\.treatment|^mu\\.time_scaled|^sigma\\.treatment|^sigma\\.time_scaled|^nu\\.time_scaled|^theta\\.time_scaled", coef_names)
]

confint(fit, parm = inference_terms)

wald_test(fit, terms = "mu.treatment")
wald_test(fit, terms = "theta.time_scaled")
```

Likelihood comparisons are useful for planned sensitivity checks when models
are nested or approximately nested and fitted to the same observed rows. Here we
compare the main model to a simpler model with linear age effects and a simpler
dependence formula.

```{r fit-reduced, results='hide'}
fit_reduced <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = margin_dist,
  copula_dist = copula_dist,
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time_scaled + age_scaled,
  sigma.formula = ~ treatment + time_scaled + age_scaled,
  nu.formula = ~ time_scaled,
  theta.formula = ~ time_scaled,
  max_outer_iter = 6,
  max_inner_iter = 6,
  outer_stop_crit = 1e-3,
  inner_stop_crit = 1e-3,
  verbose = 0
)
```

```{r likelihood-comparison}
likelihood_compare(reduced = fit_reduced, main = fit)
```

For final reporting, bootstrap inference can be used as a robustness check on
selected terms. This refits the model for each simulated dataset, so the small
`R` below is only to show the workflow; applied analyses need many more
replicates.

```{r bootstrap-run, results='hide'}
boot <- bootstrap_inference(
  fit,
  R = 1,
  terms = c("mu.treatment", "theta.time_scaled"),
  seed = 20260521,
  fit_args = list(
    max_outer_iter = 4,
    max_inner_iter = 4,
    outer_stop_crit = 1e-2,
    inner_stop_crit = 1e-2,
    use_backtracking = FALSE
  )
)
```

```{r bootstrap-output}
boot
```

## Model Checks

`check_model()` turns the main diagnostics into a compact decision object. It
does not replace visual review, but it gives a useful first pass over
convergence, calibration, tail behaviour, residual dependence, and inference
quality. The residual-dependence flag uses a practical correlation cutoff, not
a formal test; here we use `0.25` so very small remaining correlations are
reviewed visually rather than treated as automatic failures.

```{r check-model}
check <- check_model(fit, include_vcov = TRUE, dependence_cor_cutoff = 0.25)

check
check$warnings
check$recommendation
```

## Sensitivity Screens

After the first full fit, revisit the modelling choices that were shortlisted
up front. The code below refits the same regression structure across the
initial top 10 supported marginal families and all native copula families. It
is intentionally collapsed in the rendered workflow because it is a
fit-intensive sensitivity review.

<details>
<summary>Margin-copula sensitivity code</summary>

```{r margin-copula-sensitivity, eval = FALSE}
sensitivity_selection <- select_joint_distribution(
  data = dat,
  response_var = "response",
  subject_var = "subject",
  time_var = "time",
  margin_families = top_margin_families,
  copula_families = c("N", "C", "F", "G", "J", "t"),
  fit_args = list(
    mu.formula = response ~ treatment + time_scaled + s(age_scaled, bs = "ps", k = 8),
    sigma.formula = ~ treatment + time_scaled + s(age_scaled, bs = "ps", k = 5),
    nu.formula = ~ time_scaled + s(age_scaled, bs = "ps", k = 5),
    theta.formula = ~ time_scaled + s(age_scaled, bs = "ps", k = 8)
  ),
  trace = FALSE
)

head(sensitivity_selection, 15)

primary_signature <- paste(
  fit$margin_dist$family[[1]],
  fit$copula_dist,
  sep = "+"
)
primary_signature %in% paste(
  sensitivity_selection$margin_family,
  sensitivity_selection$copula_family,
  sep = "+"
)
```

</details>

Because this is a simulation, we can now compare the data-first screens with
the known generating choices. In real data these truth columns would not exist,
so this check belongs after the ordinary modelling workflow rather than before
the initial fit.

```{r truth-aware-screen}
truth_margin_rank <- margin_selection[
  margin_selection$family == "GG",
  c("family", "rank", "AIC", "delta_AIC"),
  drop = FALSE
]

truth_copula_screen <- select_copula(
  data = dat,
  u_var = "u",
  subject_var = "subject",
  time_var = "time",
  families = c("N", "C", "F", "G", "J", "t"),
  t_df_grid = c(3, 4, 6, 10, 20)
)

truth_margin_rank
truth_copula_screen
```

```{r copula-screen-overlay, fig.width = 8.75, fig.height = 6.25, out.width = "88%", fig.alt = "Binned adjacent truth uniforms with fitted copula contour overlay."}
plot_copula_fit(
  data = dat,
  copula_dist = truth_copula_screen,
  u_var = "u",
  subject_var = "subject",
  time_var = "time"
)
```

## Standard Review Checklist

For a real longitudinal regression analysis, the model review should also
include a few checks that are easy to overlook:

- confirm the intended estimand, response scale, link functions, factor
  reference levels, and time ordering before interpreting coefficients;
- review missing visits, dropout, and structurally impossible subject-time rows
  separately from ordinary missing responses;
- inspect convergence, parameter boundaries, smoothing degrees of freedom, and
  whether uncertainty estimates are stable enough for the planned inference;
- check calibration and residual patterns overall and by key groups such as
  time, treatment, age bands, and high-leverage subjects;
- check whether missingness is absent, plausibly MCAR, or associated with
  observed features that should be included in the analysis or sensitivity
  plan;
- repeat the primary conclusions under plausible marginal distributions,
  copula families, and simpler/richer formulas;
- compare against a familiar mean model when the scientific question is mostly
  about the mean, and against predictive or simulation checks when the question
  depends on tails, intervals, or trajectories;
- perform posterior- or fitted-simulation style checks against quantities that
  matter scientifically, such as trajectory ranges, event probabilities, or
  high quantiles;
- inspect influential subjects or time periods by refitting or summarising
  diagnostics after excluding obvious leverage points;
- document which warnings were resolved, which remain, and why the remaining
  limitations do not overturn the primary conclusion.

This checklist is adapted from standard regression modelling and validation
guidance, longitudinal-data texts, missing-data workflows, GAMLSS/GAM
diagnostics, and copula modelling references. In particular, Harrell (2015)
motivates pre-specified modelling strategy, validation, calibration, and
bootstrap-based uncertainty checks; Fitzmaurice, Laird, and Ware (2011) and
Diggle et al. (2002) cover repeated-measure structure, time ordering,
within-subject association, and missing longitudinal responses; van Buuren
(2018) covers practical missing-data review; Rigby and Stasinopoulos (2005)
and Wood (2017) motivate distributional and smooth-term diagnostics; and
Nelsen (2006) and Joe (2014) cover dependence, tail dependence, and copula
model assessment.

The standard model diagnostic dashboard combines PIT, QQ, worm, rootogram, and
forecast-style checks.

```{r model-diagnostics, fig.width = 8.5, fig.height = 9.5, out.width = "92%", fig.alt = "Dashboard of marginal model diagnostics including PIT, QQ, worm, rootogram, and forecast checks."}
plot(fit, data = dat)
```

## Term Diagnostics

Because this is simulated, we first plot the true term contributions used to
generate the data. The fitted term dashboard is shown immediately afterwards.

<details>
<summary>True-term plotting code</summary>

```{r true-terms, fig.width = 9.25, fig.height = 6.25, out.width = "92%", fig.alt = "Dashboard of true simulated smooth and fixed term contributions."}
term_theme <- function(p) {
  p +
    theme_minimal(base_size = 8.5) +
    theme(
      plot.title = element_text(size = 9),
      axis.title = element_text(size = 8),
      axis.text = element_text(size = 7),
      plot.caption = element_text(size = 7)
    )
}

truth_smooth_plot <- function(x, contribution, title, xlab) {
  plot_df <- data.frame(x = x, contribution = contribution)
  term_theme(
    ggplot(plot_df, aes(x = x, y = contribution)) +
      geom_line(color = "black", linewidth = 0.8) +
      labs(title = title, x = xlab, y = paste0("smooth(", xlab, ")"), caption = "truth")
  )
}

truth_continuous_plot <- function(x, contribution, title, xlab, ylab) {
  plot_df <- data.frame(x = x, contribution = contribution)
  term_theme(
    ggplot(plot_df, aes(x = x, y = contribution)) +
      geom_line(color = "black", linewidth = 0.8) +
      labs(title = title, x = xlab, y = ylab, caption = "truth")
  )
}

truth_factor_plot <- function(effects, title, xlab, ylab) {
  plot_df <- data.frame(
    x = seq_along(effects),
    level = factor(names(effects), levels = names(effects)),
    contribution = as.numeric(effects)
  )
  term_theme(
    ggplot(plot_df, aes(x = x, y = contribution)) +
      geom_hline(yintercept = 0, color = "grey70", linetype = 3) +
      geom_point(color = "black", size = 1) +
      geom_errorbar(aes(ymin = contribution, ymax = contribution), color = "red", width = 0.12) +
      scale_x_continuous(breaks = plot_df$x, labels = plot_df$level) +
      labs(title = title, x = xlab, y = ylab, caption = "truth")
  )
}

print_term_dashboard <- function(plotlist, ncol = 4) {
  nrow <- ceiling(length(plotlist) / ncol)
  grid::grid.newpage()
  grid::pushViewport(grid::viewport(layout = grid::grid.layout(nrow, ncol)))
  on.exit(grid::popViewport(), add = TRUE)
  for (i in seq_along(plotlist)) {
    row_id <- ((i - 1) %/% ncol) + 1
    col_id <- ((i - 1) %% ncol) + 1
    print(
      plotlist[[i]],
      vp = grid::viewport(layout.pos.row = row_id, layout.pos.col = col_id)
    )
  }
}

time_grid <- seq(0, 1, length.out = 200)
age_grid <- seq(0, 1, length.out = 200)

truth_term_plots <- list(
  truth_smooth_plot(age_grid, truth_mu_age_smooth(age_grid), 'mu: s(age_scaled, bs = "ps", k = 8)', "age_scaled"),
  truth_smooth_plot(age_grid, truth_sigma_age_smooth(age_grid), 'sigma: s(age_scaled, bs = "ps", k = 5)', "age_scaled"),
  truth_smooth_plot(age_grid, truth_nu_age_smooth(age_grid), 'nu: s(age_scaled, bs = "ps", k = 5)', "age_scaled"),
  truth_smooth_plot(age_grid, truth_theta_age_smooth(age_grid), 'theta: s(age_scaled, bs = "ps", k = 8)', "age_scaled"),
  truth_factor_plot(treatment_effect, "mu: treatment", "treatment", "fixed contribution: mu.treatment"),
  truth_continuous_plot(time_grid, 0.35 * time_grid, "mu: time_scaled", "time_scaled", "fixed contribution: mu.time_scaled"),
  truth_factor_plot(0.25 * c(control = 0, low = 1, high = 1), "sigma: treatment", "treatment", "fixed contribution: sigma.treatment"),
  truth_continuous_plot(time_grid, 1.25 * time_grid, "sigma: time_scaled", "time_scaled", "fixed contribution: sigma.time_scaled"),
  truth_continuous_plot(time_grid, 0.55 * time_grid, "nu: time_scaled", "time_scaled", "fixed contribution: nu.time_scaled"),
  truth_continuous_plot(time_grid, 1.80 * time_grid, "theta: time_scaled", "time_scaled", "fixed contribution: theta.time_scaled")
)

print_term_dashboard(truth_term_plots, ncol = 4)
```

</details>

```{r fitted-terms, fig.width = 8.5, fig.height = 5.5, out.width = "92%", fig.alt = "Dashboard of fitted model term effects for marginal and dependence parameters."}
plot_terms(fit, data = dat)
```

## Copula Diagnostics

Finally, inspect the fitted copula directly. The default dashboard includes
normal-scale pair diagnostics, fitted copula overlays, tail summaries, and
lagged residual checks.

```{r copula-diagnostics, fig.width = 10.5, fig.height = 10.5, out.width = "96%", fig.alt = "Copula diagnostic dashboard with residual, tail, and lag dependence checks."}
plot_copula_diagnostics(fit, data = dat)
```

## Applied Outputs

Prediction, marginal effects, and simulation answer related but different
questions. `predict()` returns marginal summaries for each requested row:
dependence has influenced the fitted coefficients, but prediction does not
condition one row on the subject's other observed responses. `simulate()` draws
full fitted-data trajectories and preserves the fitted copula dependence.
`marginal_effects()` constructs counterfactual summaries by changing a covariate
and averaging the fitted distributional parameter over the supplied data.

```{r applied-predictions}
head(predict(
  fit,
  newdata = dat,
  type = "response",
  se.fit = TRUE,
  interval = "confidence"
))

head(predict(
  fit,
  newdata = dat,
  type = "parameters"
))

head(predict(
  fit,
  newdata = dat,
  type = "quantile",
  probs = c(0.1, 0.5, 0.9)
))

head(predict(
  fit,
  newdata = dat,
  type = "cdf",
  q = median(dat$response, na.rm = TRUE)
))

head(predict(
  fit,
  newdata = dat,
  type = "density",
  y = median(dat$response, na.rm = TRUE)
))

head(predict(
  fit,
  newdata = dat,
  type = "probability",
  q = quantile(dat$response, 0.9, na.rm = TRUE),
  direction = "above"
))
```

For regression interpretation, marginal effects are often easier to communicate
than raw coefficients, especially when smooths, nonlinear links, or scale
parameters are involved. They should still be tied back to the estimand: this
summary averages over the empirical covariate distribution in `dat`.

```{r marginal-effects}
marginal_effects(
  fit,
  newdata = dat,
  variable = "treatment",
  values = levels(dat$treatment),
  parameter = "mu",
  se.fit = TRUE
)
```

Fitted-model simulation is useful for predictive checks, planning, and
uncertainty communication because it simulates both the marginal distribution
and the within-subject copula dependence.

```{r fitted-simulation}
sim_model <- simulate(fit, nsim = 3, seed = 20260521)

head(sim_model)
summary(sim_model)
```

## Benchmark Against Familiar Models

Benchmarking helps clarify whether a longitudinal GAMLSS-copula model is adding
value for the question at hand. If the estimand is only the mean and the
response is close to a standard distribution, simpler mean models may be
adequate. The distributional model is most useful when the scientific question
depends on tails, quantiles, heterogeneity, uncertainty, or within-subject
dependence.

```{r benchmark}
bench <- benchmark_standard_models(
  data = dat,
  formula = response ~ treatment + time_scaled + s(age_scaled, bs = "ps"),
  subject_var = "subject",
  family = "gamma",
  comparators = "gam",
  fit = fit,
  fit_name = "gamlss_longitudinal"
)

bench
```

## References

Diggle, P. J., Heagerty, P., Liang, K.-Y., and Zeger, S. L. (2002).
*Analysis of Longitudinal Data*. 2nd ed. Oxford University Press.
<https://academic.oup.com/book/54026>.

Fitzmaurice, G. M., Laird, N. M., and Ware, J. H. (2011). *Applied
Longitudinal Analysis*. 2nd ed. John Wiley & Sons.
<https://www.wiley-vch.de/en/areas-interest/mathematics-statistics/applied-longitudinal-analysis-978-0-470-38027-7>.

Harrell, F. E. Jr. (2015). *Regression Modeling Strategies: With Applications
to Linear Models, Logistic and Ordinal Regression, and Survival Analysis*. 2nd
ed. Springer. <https://doi.org/10.1007/978-3-319-19425-7>.

Joe, H. (2014). *Dependence Modeling with Copulas*. Chapman & Hall/CRC.
<https://copula.stat.ubc.ca/>.

Nelsen, R. B. (2006). *An Introduction to Copulas*. 2nd ed. Springer.
<https://doi.org/10.1007/0-387-28678-0>.

Rigby, R. A. and Stasinopoulos, D. M. (2005). Generalized additive models for
location, scale and shape. *Journal of the Royal Statistical Society: Series C
(Applied Statistics)*, 54(3), 507-554.
<https://doi.org/10.1111/j.1467-9876.2005.00510.x>.

van Buuren, S. (2018). *Flexible Imputation of Missing Data*. 2nd ed. Chapman
& Hall/CRC. <https://stefvanbuuren.name/fimd/>.

Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R*. 2nd
ed. Chapman & Hall/CRC.
<https://www.routledge.com/Generalized-Additive-Models-An-Introduction-with-R-Second-Edition/Wood/p/book/9781498728331>.

## Next Steps

In a real analysis, the same workflow should be repeated as sensitivity checks
when conclusions depend on modelling choices: revisit the top marginal
families, plausible copula families, simpler versus richer formulas, and
benchmark models. For simulation studies, the stored `true_*` columns make it
straightforward to compare fitted effects against the data-generating truth. For
applied studies, report the model only after the diagnostic warnings are either
resolved or clearly described as limitations.
