---
title: "Non-Gaussian Longitudinal Outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-Gaussian Longitudinal Outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide is for longitudinal outcomes where a Gaussian working model is too
thin: positive skewed responses, counts, zero-heavy responses, bounded
measurements, or outcomes where tail probabilities matter. These are the cases
where `gamlss.longitudinal` should feel like a natural first model rather than
a specialist fallback.

## Why Not Just A GEE Or GLMM?

GEEs and GLMMs can be excellent mean models. They are less direct when the
scientific question is distributional:

- What is the 90th percentile under treatment?
- Does treatment change variability, not just the mean?
- Are upper-tail events clustered within subjects?
- Does a Poisson or Gamma mean model hide dispersion or asymmetry?
- Are predicted probabilities calibrated across the response range?

For these questions, start by choosing a response distribution, then model its
parameters and the longitudinal dependence.

## Positive Continuous Outcomes

For positive skewed outcomes, start with Gamma, inverse Gaussian, Weibull, or
other supported positive margins.

```{r positive-screen}
screen_margin(
  dat,
  response_var = "response",
  candidate_families = c("GA", "IG", "WEI", "LOGNO")
)
```

Then fit a distributional model. Put the main mean-scale question in `mu`, and
use `sigma` when variability changes by time, treatment, or baseline risk.

```{r positive-fit}
fit_pos <- fit_longitudinal(
  dataset = dat,
  margin_dist = gamlss.dist::GA(mu.link = "log", sigma.link = "log"),
  copula_dist = "N",
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + age_scaled,
  sigma.formula = ~ treatment + time,
  theta.formula = ~ time
)

check_model(fit_pos)$checks
```

Report response-scale quantities:

```{r positive-answers}
predict(fit_pos, type = "response", se.fit = TRUE, interval = "confidence")
predict(fit_pos, type = "quantile", probs = c(0.1, 0.5, 0.9))
predict(fit_pos, type = "probability", q = 10, direction = "above")

marginal_effects(
  fit_pos,
  newdata = dat,
  variable = "treatment",
  parameter = "mu",
  se.fit = TRUE
)
```

## Count Outcomes

For counts, begin with Poisson only if the count process is plausibly simple.
Move to overdispersed or zero-adjusted families when screening and diagnostics
support it.

```{r count-screen}
screen_margin(
  count_dat,
  response_var = "response",
  candidate_families = c("PO", "NBI", "NBII", "ZIP", "ZINBI")
)
```

```{r count-fit}
fit_count <- fit_longitudinal(
  dataset = count_dat,
  margin_dist = gamlss.dist::NBI(),
  copula_dist = "N",
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + age_scaled,
  sigma.formula = ~ treatment,
  theta.formula = ~ 1
)

check_model(fit_count)$tail
```

For applied reporting, avoid stopping at log-link coefficients. Give predicted
means, selected probabilities, upper-tail risks, or simulated trajectories.

```{r count-answers}
predict(fit_count, type = "response")
predict(fit_count, type = "quantile", probs = c(0.5, 0.9, 0.95))
simulate(fit_count, nsim = 100, type = "copula")
```

## Diagnostics That Matter More For Non-Gaussian Outcomes

Use `check_model()` as the first diagnostic pass:

```{r diagnostics}
check <- check_model(fit_pos)
check$checks
check$pit
check$tail
check$residual_dependence
```

Read the warnings literally:

- Marginal calibration warnings mean the chosen response family is still wrong.
- Tail warnings mean the model is missing high or low outcome risk.
- Residual dependence warnings mean the copula is not absorbing the repeated
  measurement association.
- Convergence or Hessian warnings mean inference should be described as
  approximate or revisited with simpler formulas.

For a fuller checklist of coefficient, response-mean, quantile, and simulation
uncertainty language, see `vignette("inference-uncertainty", package =
"gamlss.longitudinal")`.

## Benchmark The Claim

Use adoption benchmarks when claiming that the distributional model improves on
standard mean-model practice.

```{r benchmark}
positive_bench <- run_adoption_benchmarks(
  scenarios = "gamma_positive",
  reps = 50,
  methods = c("rs_separate", "gee", "glmm", "gam"),
  max_elapsed_sec = 60,
  write_results = TRUE
)

positive_bench$summary$summary
```

For counts:

```{r count-benchmark}
count_bench <- run_adoption_benchmarks(
  scenarios = "poisson_count",
  reps = 50,
  methods = c("rs_separate", "gee", "glmm", "gam"),
  max_elapsed_sec = 60,
  write_results = TRUE
)

count_bench$summary$summary
```

Use the benchmark summary to make narrow claims: mean prediction, interval
calibration, 90th percentile error, upper-tail behaviour, convergence, or
runtime. A model that wins on the mean but loses on tails is not a
distributional replacement.

For non-Gaussian comparator rows, the most useful first checks are:

- `benchmark_q90_mae`: how far the model-implied 90th percentile is from the
  simulated truth.
- `benchmark_upper_tail_error_90`: whether the model over- or under-predicts
  probability above the true 90th percentile.
- `benchmark_mean_rmse`: whether the response mean is still competitive.

## When To Keep The Simpler Model

Keep a GEE or GLMM as the primary model when the response family is routine,
diagnostics are clean, the scientific estimand is only a marginal or
subject-specific mean, and quantiles or tail probabilities would not change the
decision. In that setting, `gamlss.longitudinal` is still a useful sensitivity
analysis for distributional assumptions.
