---
title: "Replacing GEE and GLMM Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replacing GEE and GLMM Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide is for analysts who would usually start with a GEE or GLMM. The goal
is not to claim that those models are obsolete in every setting. The goal is to
make the GAMLSS-copula workflow the default when the response distribution,
heteroskedasticity, tails, asymmetry, or time-varying dependence matter.

For focused migration examples, see the companion guides
`vignette("replace-a-gee", package = "gamlss.longitudinal")` and
`vignette("replace-a-random-intercept-glmm", package = "gamlss.longitudinal")`.
For the default end-to-end workflow, see
`vignette("standard-workflow", package = "gamlss.longitudinal")`.
For a primary-model versus sensitivity-model decision rubric, see
`vignette("adoption-decision-guide", package = "gamlss.longitudinal")`.
For distribution-first analyses, see
`vignette("non-gaussian-longitudinal-outcomes", package = "gamlss.longitudinal")`.
For incomplete panels, see `vignette("missing-visits", package =
"gamlss.longitudinal")`.
For changing longitudinal association, see
`vignette("time-varying-dependence", package = "gamlss.longitudinal")`.
For simulation evidence, see
`vignette("benchmarking-adoption", package = "gamlss.longitudinal")`.
For reporting intervals and uncertainty, see
`vignette("inference-uncertainty", package = "gamlss.longitudinal")`.
For diagnostic decision rules, see
`vignette("diagnostics-decisions", package = "gamlss.longitudinal")`.

## Translation Table

| GEE/GLMM habit | `gamlss.longitudinal` equivalent |
|---|---|
| Pick a mean model | Model `mu.formula` |
| Pick a variance/random-effect structure | Model `sigma.formula` and copula dependence |
| Choose a working correlation | Use `select_copula()` and `theta.formula` |
| Check residual plots | Use `check_model()`, `plot()`, `plot_terms()`, and `plot_copula_diagnostics()` |
| Interpret fixed effects | Use `summary()`, `marginal_effects()`, or `effects()` |
| Predict means or intervals | Use `predict(type = "response")` and `predict(type = "quantile")` |
| Simulate fitted trajectories | Use `simulate(type = "copula")` |

## Standard Workflow

Start by screening the marginal response. This is deliberately marginal: it is
only a guide before the longitudinal dependence model is fitted.

```{r margin-screen}
margin_screen <- screen_margin(dat, response_var = "response")
margin_screen
```

Then screen copula families from pseudo-observations. In real analyses these
usually come from a preliminary marginal model; in simulations they may be
stored directly.

```{r copula-screen}
copula_screen <- select_copula(
  data = dat,
  u_var = "u",
  subject_var = "subject",
  time_var = "time",
  families = c("N", "C", "F", "G", "J", "t")
)
copula_screen
```

Fit the longitudinal model with distributional formulas using
`gamlss_longitudinal()`. Keep copula screening and model checking explicit:
use `select_copula()` before fitting and `check_model()` after fitting.
A typical first model puts the main scientific effects in `mu`, obvious
heteroskedasticity in `sigma`, and time or subject-level drivers of dependence
in `theta`.

```{r fit}
fit <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = gamlss.dist::GA(mu.link = "log", sigma.link = "log"),
  copula_dist = attr(copula_screen, "selected"),
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + s(age_scaled, bs = "ps"),
  sigma.formula = ~ treatment + time,
  theta.formula = ~ time
)
```

Use `check_model()` as the first diagnostic pass. It returns numeric summaries,
plain-language decisions, tail calibration summaries, and optional plot
objects.

```{r check}
check <- check_model(fit)
check
```

For applied interpretation, prefer predictions and counterfactual summaries over
raw coefficient inspection.

```{r interpretation}
predict(fit, type = "quantile", probs = c(0.1, 0.5, 0.9))
predict(fit, type = "response", se.fit = TRUE, interval = "confidence")

marginal_effects(
  fit,
  newdata = dat,
  variable = "treatment",
  parameter = "mu",
  se.fit = TRUE
)
```

## Inference Language

For a fuller guide to reporting uncertainty, see
`vignette("inference-uncertainty", package = "gamlss.longitudinal")`.

For coefficient intervals, use `confint(fit)`. The reported intervals depend on
the variance-covariance route requested through `vcov()`: analytical Hessian
when available, numerical Hessian when requested or used as a fallback, and
cached fit-time covariance when present. When numerical Hessians are used,
describe the intervals as approximate numerical-Hessian likelihood intervals.

For coefficient hypotheses, use `wald_test(fit, terms = ...)` or a custom
contrast matrix. These tests use the same variance-covariance route as
`summary()` and `confint()`, so the same Hessian and fallback caveats apply.
For nested model comparisons, use `likelihood_compare(reduced, full)` and state
that the comparison uses the fitted joint likelihood.

Prediction intervals from `predict(..., se.fit = TRUE, interval = "confidence")`
are delta-method intervals for the fitted response mean. They are designed for
fast applied summaries, not as a replacement for bootstrap or simulation-based
uncertainty when the full trajectory distribution is the target.

## Benchmarking Against Familiar Models

For a quick sensitivity comparison, `benchmark_standard_models()` fits common
baselines when their optional packages are installed. The result is not a full
simulation study by itself; it is a standardized fitting record that can feed
coverage, calibration, and tail-behavior benchmarks.

```{r benchmark-standard-models}
bench <- benchmark_standard_models(
  data = dat,
  formula = response ~ treatment + time + age_scaled,
  subject_var = "subject",
  family = "gaussian",
  comparators = c("gee", "glmm", "gam")
)

bench
```

Use the benchmark table to separate different questions: GEEs estimate marginal
mean effects, GLMMs estimate conditional mean effects, GAMs focus on smooth
mean structure, and `gamlss.longitudinal` models the response distribution and
dependence jointly. That distinction matters when claiming a win, tie, or loss.

For repeated simulation evidence, request these comparators in
`run_coverage_simulations()`:

```{r benchmark-coverage}
results <- run_coverage_simulations(
  families = "NO",
  copulas = "N",
  methods = c("rs_separate", "gee", "glmm", "gam"),
  designs = "covariate",
  write_results = FALSE
)
```

In simulated runs, `benchmark_mean_bias`, `benchmark_mean_mae`, and
`benchmark_mean_rmse` compare each standard model's fitted mean against the
simulated `true_mu`. Gaussian comparator rows also include residual-calibrated
PIT, tail, and 95% interval diagnostics. Those are useful screens for
misspecification, while formal coverage claims should still come from repeated
simulation summaries.

```{r benchmark-summary}
bench_summary <- summarise_benchmark_results(
  results,
  metrics = c("benchmark_mean_rmse", "benchmark_interval_coverage_95", "elapsed_sec")
)

bench_summary
```

The summary labels each method as a win, tie, loss, or missing within each
simulation case, then aggregates those labels by metric. Lower error and runtime
are better, interval coverage targets 0.95, and calibration/tail errors target
0.

## When Not To Use This Package

Use a simpler GEE or GLMM when the scientific question is only about a marginal
mean, the response distribution is well handled by a standard family, the
dependence model is nuisance-only, and the added distributional complexity would
not change decisions.

Be cautious when the number of subjects is very small, the time grid is sparse
or irregular in ways not represented by adjacent-time copulas, the chosen
distribution is weakly identified, or the model requires random effects that are
the scientific target rather than a device for dependence.

In those cases, `gamlss.longitudinal` can still be useful as a sensitivity
analysis, but it should not be oversold as an automatic replacement.
