---
title: "Replace a GEE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replace a GEE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide is for an analyst who would usually reach for
`geepack::geeglm()`: a longitudinal marginal mean model with a working
correlation. `gamlss.longitudinal` is a good replacement candidate when the
mean is not the only scientific object, or when the response distribution,
heteroskedasticity, tails, asymmetry, or time-varying dependence can affect the
answer.

## Translation

| GEE step | `gamlss.longitudinal` step |
|---|---|
| Choose a mean formula | Put the same scientific predictors in `mu.formula` |
| Choose a working correlation | Screen copulas with `select_copula()` |
| Treat variance as nuisance | Model `sigma.formula` when spread changes over time or treatment |
| Report robust mean effects | Report `summary()`, `confint()`, `wald_test()`, and `marginal_effects()` with inference method stated |
| Check residuals informally | Use `check_model()` decisions plus PIT, tail, and dependence diagnostics |
| Predict the mean | Use `predict(type = "response")`; add quantiles when distributional answers matter |

## Typical GEE

```{r gee-fit}
gee_fit <- geepack::geeglm(
  response ~ treatment + time + age_scaled,
  id = subject,
  data = dat,
  family = gaussian(),
  corstr = "exchangeable"
)
```

That model answers a marginal mean question. It does not directly model changing
scale, tail behavior, or a likelihood for subject trajectories.

## GAMLSS-Copula Replacement

Start with the same mean structure in `mu.formula`, then add only the
distributional and dependence structure that has a scientific or diagnostic
reason to be there.

```{r screen}
screen_margin(dat, response_var = "response")

copula_screen <- select_copula(
  data = dat,
  u_var = "u",
  subject_var = "subject",
  time_var = "time"
)
```

```{r fit}
fit <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = gamlss.dist::NO(),
  copula_dist = attr(copula_screen, "selected"),
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + age_scaled,
  sigma.formula = ~ treatment + time,
  theta.formula = ~ time
)
```

## Diagnostics To Decide Whether The Replacement Helped

```{r check}
check <- check_model(fit)
check$checks
check$tail
check$residual_dependence
```

Read these checks as decisions:

- Marginal calibration warnings mean the response distribution is still off.
- Tail warnings mean the fitted distribution under- or over-represents extreme
  outcomes.
- Residual dependence warnings mean the copula did not absorb the longitudinal
  association.

## Applied Answers

For a GEE-like mean answer:

```{r mean-answer}
predict(fit, type = "response", se.fit = TRUE, interval = "confidence")
wald_test(fit, terms = "mu.treatment", method = "analytical")

marginal_effects(
  fit,
  newdata = dat,
  variable = "treatment",
  parameter = "mu",
  se.fit = TRUE
)
```

For answers a GEE usually cannot give directly:

```{r distributional-answer}
predict(fit, type = "quantile", probs = c(0.1, 0.5, 0.9))
predict(fit, type = "probability", q = 10, direction = "above")
simulate(fit, nsim = 100, type = "copula")
```

## Sensitivity Benchmark

A replacement claim should be checked, not assumed. Fit the familiar GEE next
to the GAMLSS-copula model and summarize the repeated-simulation evidence.

```{r benchmark}
results <- run_coverage_simulations(
  families = "NO",
  copulas = "N",
  methods = c("rs_separate", "gee", "gam"),
  designs = "covariate",
  write_results = FALSE
)

summarise_benchmark_results(
  results,
  metrics = c("benchmark_mean_rmse", "benchmark_interval_coverage_95", "elapsed_sec")
)
```

## When To Keep The GEE

Keep a GEE as the primary analysis when the scientific estimand is only a
marginal mean, the response family is routine, dependence is nuisance-only, and
diagnostics do not suggest distributional or tail misspecification. In that
setting, `gamlss.longitudinal` is still useful as a sensitivity analysis, but a
simple GEE may be the cleaner primary model.
