---
title: "Replace a Random-Intercept GLMM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replace a Random-Intercept GLMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide is for an analyst who would usually start with
`lme4::lmer()` or `lme4::glmer()` using a subject-level random intercept.
`gamlss.longitudinal` is a good replacement candidate when the random
intercept is mainly a device for within-subject association, and the applied
question also needs flexible margins, heteroskedasticity, tail behavior,
asymmetry, or time-varying dependence.

## Translation

| Random-intercept GLMM habit | `gamlss.longitudinal` step |
|---|---|
| Fixed-effect mean formula | Put the same scientific predictors in `mu.formula` |
| `(1 | subject)` for repeated measures | Model repeated-measure dependence with a copula |
| Extra dispersion or residual spread | Model `sigma.formula` or another distribution parameter |
| Subject-specific fitted values | Use fitted response predictions or `simulate(type = "copula")` for trajectories |
| Random-effect variance as dependence summary | Use copula parameter summaries and `theta.formula` |
| GLMM residual checks | Use `check_model()` for marginal, tail, and residual-dependence decisions |

## Typical Random-Intercept GLMM

For a continuous response:

```{r lmer-fit}
glmm_fit <- lme4::lmer(
  response ~ treatment + time + age_scaled + (1 | subject),
  data = dat
)
```

For a count response:

```{r glmer-fit}
count_glmm_fit <- lme4::glmer(
  response ~ treatment + time + age_scaled + (1 | subject),
  data = dat,
  family = poisson()
)
```

These models are often strong defaults when the conditional mean model and
subject heterogeneity are the scientific focus. They are less direct when the
response distribution, changing scale, tail clustering, or interpretable
time-varying dependence is central to the analysis.

## GAMLSS-Copula Replacement

Start by preserving the fixed-effect scientific question in `mu.formula`.
Then replace the random intercept with a copula dependence model. A constant
`theta.formula = ~ 1` is the closest analogue to a simple random-intercept
dependence assumption; `theta.formula = ~ time` is useful when dependence may
change over follow-up.

```{r screen}
screen_margin(dat, response_var = "response")

copula_screen <- select_copula(
  data = dat,
  u_var = "u",
  subject_var = "subject",
  time_var = "time"
)
```

```{r fit}
fit <- fit_longitudinal(
  dataset = dat,
  margin_dist = gamlss.dist::NO(),
  copula_dist = attr(copula_screen, "selected"),
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + age_scaled,
  sigma.formula = ~ treatment + time,
  theta.formula = ~ 1
)
```

For a count response, choose a supported count margin during screening and fit
that family instead:

```{r count-fit}
count_fit <- fit_longitudinal(
  dataset = count_dat,
  margin_dist = gamlss.dist::PO(),
  copula_dist = "N",
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + age_scaled,
  theta.formula = ~ 1
)
```

## Interpret The Estimand

A GLMM coefficient is usually conditional on the latent random effect.
The GAMLSS-copula replacement describes distributional parameters and
within-subject dependence without requiring a latent subject intercept as the
main explanation. That can be an advantage when the goal is a response-scale
distributional answer, but it should be stated clearly in reports.

```{r interpretation}
confint(fit)

marginal_effects(
  fit,
  newdata = dat,
  variable = "treatment",
  parameter = "mu",
  se.fit = TRUE
)

predict(fit, type = "response", se.fit = TRUE, interval = "confidence")
predict(fit, type = "quantile", probs = c(0.1, 0.5, 0.9))
simulate(fit, nsim = 100, type = "copula")
```

## Diagnostics To Decide Whether The Replacement Helped

```{r check}
check <- check_model(fit)
check$checks
check$tail
check$residual_dependence
check$copula
copula_time_summary(fit)
```

Use these diagnostics to separate three common outcomes:

- If marginal calibration improves but residual dependence remains, keep the
  margin and revisit `theta.formula` or the copula family.
- If dependence is well captured but tail checks fail, revisit the response
  distribution before adding more mean predictors.
- If the GAMLSS-copula model and GLMM give the same answer with clean
  diagnostics, the simpler GLMM may be easier to defend.

## Sensitivity Benchmark

For replacement claims, compare the GAMLSS-copula model to the familiar GLMM
inside repeated simulations or an applied sensitivity table.

```{r benchmark}
results <- run_coverage_simulations(
  families = "NO",
  copulas = "N",
  methods = c("rs_separate", "glmm", "gam"),
  designs = "covariate",
  write_results = FALSE
)

summarise_benchmark_results(
  results,
  metrics = c("benchmark_mean_rmse", "benchmark_interval_coverage_95", "elapsed_sec")
)
```

## When To Keep The GLMM

Keep the GLMM as the primary analysis when subject-specific latent heterogeneity
is the scientific target, when the response family and variance assumptions are
adequate, or when the model needs random slopes or hierarchical structure that
is not naturally expressed as a copula dependence model. In those settings,
`gamlss.longitudinal` can still be valuable as a sensitivity analysis focused
on distributional assumptions and tail behavior.
