---
title: "Simulator usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulator usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette shows how to generate longitudinal datasets with GAMLSS margins,
first-order copula dependence, covariates, stored truth columns, and optional
missing responses. The simulator is intended for examples, method checks,
benchmark scenarios, and teaching datasets where the data-generating mechanism
should be explicit.

```{r packages}
library(gamlss.longitudinal)
library(gamlss.dist)
```

## Minimal Data

`simulate_longitudinal_dataset()` returns one row per subject-time observation.
The marginal distribution is supplied as a `gamlss.dist` family object, and the
copula is supplied by family code:

- `"N"` Gaussian
- `"C"` Clayton
- `"F"` Frank
- `"G"` Gumbel
- `"J"` Joe
- `"t"` Student t

```{r minimal-data}
dat <- simulate_longitudinal_dataset(
  n = 100,
  times = 0:4,
  margin_dist = NO(),
  margin_params = list(mu = 0, sigma = 1),
  copula_dist = "N",
  copula_params = list(theta = 0.35),
  seed = 1
)

head(dat)
table(dat$time)
```

By default the output includes `u`, the simulated marginal uniforms, plus
`true_*` columns for every marginal and copula parameter. The first time point
has no incoming adjacent-time copula edge, so `true_theta` is `NA` there.

## Time-Varying Parameters

Parameter specifications can be constants, vectors indexed by time, matrices
indexed by subject and time, or functions of the generated long-format design.
Vectors for marginal parameters should have one value per time point. Vectors
for copula parameters should have one value per adjacent time edge.

```{r time-varying-parameters}
dat <- simulate_longitudinal_dataset(
  n = 80,
  times = 1:4,
  margin_dist = GA(mu.link = "log", sigma.link = "log"),
  margin_params = list(
    mu = function(d) exp(1 + 0.15 * d$.sim_time_index),
    sigma = c(0.35, 0.45, 0.55, 0.65)
  ),
  copula_dist = "C",
  copula_params = list(tau = c(0.15, 0.25, 0.35)),
  seed = 2
)

head(dat)
aggregate(true_sigma ~ time, dat, unique)
aggregate(true_theta ~ time, dat, unique, na.rm = TRUE)
```

The copula parameter can be supplied as `theta`/`par`, or as Kendall's `tau`.
Using `tau` is often easier when the goal is to describe dependence strength on
a common scale across copula families.

## Covariates

Use `simulate_longitudinal_covariates()` inside the `covariates` callback when
you want a clear split between subject-level variables and observation-level
variables. Subject-level variables are generated once per subject and repeated
across time. Observation-level variables are generated once per row.

```{r covariates}
treatment_effect <- c(control = 0, active = 0.4)

dat <- simulate_longitudinal_dataset(
  n = 120,
  times = seq(0, 1, length.out = 5),
  margin_dist = GA(mu.link = "log", sigma.link = "log"),
  covariates = function(base) {
    simulate_longitudinal_covariates(
      base,
      subject = list(
        treatment = function(s) {
          factor(
            sample(c("control", "active"), nrow(s), replace = TRUE),
            levels = c("control", "active")
          )
        },
        age = function(s) rnorm(nrow(s), mean = 55, sd = 10)
      ),
      observation = list(
        time_scaled = function(d) sim_rescale01(d$time)
      )
    )
  },
  margin_params = list(
    mu = function(d) {
      age_scaled <- sim_rescale01(d$age)
      eta <- 1.2 +
        sim_factor_effect(d$treatment, treatment_effect) +
        0.35 * d$time_scaled +
        sim_smooth_bump(age_scaled, amplitude = 0.25)
      exp(eta)
    },
    sigma = function(d) {
      exp(-0.8 + 0.25 * d$time_scaled)
    }
  ),
  copula_dist = "N",
  copula_params = list(
    theta = function(e) {
      0.2 + 0.25 * sim_rescale01(e$time_left)
    }
  ),
  seed = 3
)

head(dat)
```

Inside marginal parameter functions, `d` is the long-format subject-time data.
Inside copula parameter functions, `e` is the adjacent-pair edge data. Edge data
contains the covariates from the left and right rows where needed, including
time variables such as `time_left` and `time_right`.

The helper functions `sim_rescale01()`, `sim_factor_effect()`, and the
`sim_smooth_*()` shapes are small building blocks for readable truth
specifications. They are optional; ordinary R code works as well.

## Missing Responses

The simulator creates complete balanced panels. Missing responses can be added
after simulation by setting the response column to `NA`. This keeps the
missingness mechanism visible in the vignette or benchmark code.

```{r missing-responses}
set.seed(4)

missing_probability <- plogis(
  -1.5 +
    0.8 * dat$time_scaled +
    0.4 * (dat$treatment == "active")
)

missing_rows <- rbinom(nrow(dat), size = 1, prob = missing_probability) == 1
dat$response[missing_rows] <- NA_real_
dat$response_missing <- is.na(dat$response)

mean(dat$response_missing)

check_missingness(
  dat,
  response_var = "response",
  time_var = "time",
  subject_var = "subject",
  predictors = c("time_scaled", "treatment", "age")
)
```

Missing responses are allowed by `gamlss_longitudinal()`, but predictor values
in submitted rows must be observed. Structural missing visits are handled by
the fitter through its subject-time grid expansion; for simulation studies it
is usually clearer to record whether a scenario uses response missingness,
dropped visits, or both.

## Fitting The Simulated Data

Because the simulator can store the true parameter values, it is useful for
checking whether a model recovers the main data-generating features.

```{r fit-simulated-data}
fit <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = GA(mu.link = "log", sigma.link = "log"),
  copula_dist = "N",
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time_scaled + s(age, bs = "ps"),
  sigma.formula = ~ time_scaled,
  theta.formula = ~ time_scaled
)

summary(fit)
check_model(fit)
plot(fit)
plot_copula_diagnostics(fit, data = dat)
```

For an end-to-end worked example with diagnostics, interpretation, prediction,
and benchmark comparisons, see
`vignette("native-simulation-workflow", package = "gamlss.longitudinal")`.
