---
title: "Standard workflow: screening, fit, inference, diagnostics, and simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard Workflow API, Inference, and Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette gives the minimal standard workflow for applied longitudinal analyses with `gamlss.longitudinal`.

We include a suggested set of steps and functions including short descriptions of each, while excluding outputs inline to minimise the complexity of reviewing the file. The idea is to allow the user to quickly pickup the relevant code and apply to either their own dataset or the included simulated dataset.

For a worked through example with more detailed review of outputs and diagnostics, including their interpretation, please review the detailed worked example: [Detailed worked example](native-simulation-workflow.html).

The key parts are:

1. Select marginal distribution and copula function
2. Fit model, review outputs and compare fits
3. Review model diagnostics
4. (Optional) Predict and simulate
5. (Optional) Compare to common models: gee, glmm, gam

1-3 are the core model building workflow, while 4-5 provide additional tools for analaysis depening on user needs.

## Installation

For now installation is available directly from the project's github repository. You may need to install the package 'remotes' through `install.packages("remotes")`.

```{r installation}
remotes::install_github("ahibbert/gamlss.longitudinal")
library(gamlss.longitudinal)
```

## Data Requirements

The dataset should be structured as one row per subject and timepoint. The dataset should include:

- the response variable;
- a subject identifier;
- a visit, time, or occasion identifier;
- any covariates for analysis.

This workflow is designed to be able to be used on a dataset of your choice (with the suggested structure). Alternatively, the small simulated dataset below has the required structure and can be used to run the workflow end to end.

```{r simulated-data}
# Simulate example dataset
dat <- simulate_longitudinal_dataset(
  n = 100,
  times = 0:3,
  margin_dist = GG(),
  copula_dist = "N",
  margin_params = list(
    mu = function(d) exp(1.6 + 0.35 * (d$treatment == "active") + 0.12 * d$time + d$subject_shift),
    sigma = function(d) exp(-0.5 + 0.08 * d$time)
  ),
  copula_params = list(theta = 0.4),
  covariates = function(d) {
    simulate_longitudinal_covariates(d, subject = list(
      treatment = function(s) factor(rep(c("control", "active"), length.out = nrow(s))),
      age_scaled = function(s) as.numeric(scale(rnorm(nrow(s), mean = 55, sd = 10))),
      subject_shift = function(s) rnorm(nrow(s), sd = 0.35)
    ))
  },
  seed = 1
)
```

```{r data-shape}
# Review data completeness / counts by time, subject
str(dat)
table(dat$subject)
table(dat$time)
```

Missing responses are allowed but not directly accounted for as part of model fit so the missingness mechanism
still needs to be reviewed as part of model assessment ideally. Missingness at random is generally well managed by the model, but if there is missingness not at random this may affect results. See
`vignette("missing-visits", package = "gamlss.longitudinal")` for that case.

## Step 1: Select margin distribution and copula function

First, we review the shape of the response distribution (margin) and choose the best fitting distribution. **gamlss_longitudinal** supports all non-mixed response types from the **gamlss.dist** library of distributions.

For simplicty, we provide the function `select_margin()` which fits the available gamlss distributions and provides a sorted list of fits by AIC, and confirms if they are supported by gamlss_longitudinal. This can be limited to certain fits such as specific distributions, e.g. `families = c("PO", "GA")` or response types, e.g. `type=realplus`.

In general, we would suggest using the top fitting distributions as a shortlist that can later be revisited once covariates are incorporated into the fit.

```{r margin-screen}
# Select best-fitting margin distribution
margin_selection <- select_margin(
  dat,
  response_var = "response"
)

print(margin_selection)

margin_dist <- best_fit_family(margin_selection)
```

We then select our starting copula based on the chosen marginal distribution. The function `select_copula()` is used to compare dependence copula functions. When `response_var` and `margin_dist` are supplied, pseudo-observations are created automatically from a temporary marginal GAMLSS fit.

```{r copula-screen}
# Select best-fitting copula distribution
copula_selection <- select_copula(
  data = dat,
  response_var = "response",
  margin_dist = margin_dist,
  subject_var = "subject",
  time_var = "time"
)

print(copula_selection)

copula_dist <- best_fit_family(copula_selection)
```

The selected family is a starting point. Dependence diagnostics after the full fit matter more than the screen as dependence shape can change a lot after covariates are accounted for in the margins.

## Step 2: Fit the longitudinal gamlss model and review key outputs

Depending on the margin distribution selected and the copula function selected, it's possible to depend variables on any parameter for the margin: `mu.formula` which is generally the mean, `sigma.formula` which is generally the variance parameter, `nu.formula`, `tau.formula`, and for the copula: `theta.formula` and `zeta.formula`.

Other than the covariate formulas, the model requires the user specify the subject and time variables in the provided dataset through parameters `time_var` and `subject_var`. The user should also specify the chosen margin distribution in functional form (i.e. `GA()`) as well as the selected distributions for the margin and copula (initially based on screening in the previous step).

For the included simulated example, we know there is a binary treatment effect and a time effect on both the mean, `mu.formula`, and the variance, `sigma.formula`. In addition, there is a smooth age-effect on the mean. There is also increasing correlation over time which we capture with a covariate in `theta.formula`.

```{r fit}
# Fit longitudinal gamlss model
fit <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = margin_dist,
  copula_dist = copula_dist,
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + s(age_scaled, bs = "ps"),
  sigma.formula = ~ treatment + time,
  theta.formula = ~ time
)
```

You can then review a summary of the fit including covariate estimates, standard errors and significance alongside model selection criteria: AIC, BIC and log likelihood.

```{r final-vcov}
# Model summary
summary(fit)
```

Other standard functions for outputs of regression models are also available, such as confidence intervals, wald tests, likelihood ratios and bootstrap inference.

```{r final-postfit-tests}
# Confidence intervals
confint(fit)

#Wald test
wald_test(fit)

# Likelihood ratio test
fit_reduced <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = margin_dist,
  copula_dist = copula_dist,
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time,
  sigma.formula = ~ treatment + time,
  theta.formula = ~ time
)
likelihood_compare(fit_reduced, fit)

# Bootstrapped confidence intervals (WARNING: Can take a long time to run since it's refitting for each R)
boot <- bootstrap_inference(fit, R = 20)
print(boot)
```

See `vignette("inference-uncertainty", package = "gamlss.longitudinal")` for
the reporting language around Hessian, numerical-Hessian, delta-method, and
simulation-based uncertainty.

See `vignette("diagnostics-decisions", package = "gamlss.longitudinal")` for
the decision language around marginal, tail, dependence, and inference warnings.

## Step 3: Diagnostics

We provide a built-in check_model() function which highlights any warnings and recommendations in plain language from the fit diagnostics incuding convergence errors or unusual looking fit statistics.

```{r check}
# Model checks
check <- check_model(fit, include_vcov = TRUE)
check
check$warnings
check$recommendation
```

To visually review the fits for both margin and the dependence copula we provide plotting functions directly on the fit object: `plot(fit)` and `plot_copula_diagnostics(fit)`. For the margin plots, we include standard margin diagnostics:

- PIT histogram,
- QQ plot,
- Worm plot, and
- Rootogram.

For the copula we include empirical copula v fitted density and correlation by quantile, Rosenblatt residuals and lag plot, ordered fitted versus observed density, tail co-occurrence and exceedence, and residual dependence by lag.

```{r plot chekc}
# View margin diagnostics
plot(fit)

# View dependence diagnostics
plot_copula_diagnostics(fit)


```

Treat the output as a decision aid:
- convergence warnings mean either not enough iterations have been allowed, or model is misspecified
- marginal warnings mean the response family or parameter formulas may need adjustment;
- tail warnings mean quantile or probability claims are not yet trustworthy;
- residual-dependence warnings mean the copula family or `theta.formula` / `zeta.formula` may be misspecified;
- inference warnings mean intervals should be reported as approximate or the model should be simplified.

Ideally all warnings should be rectified and diagnostics accepted as satisfactory prior to moving to reporting.

## Optional: Simulation and prediction

Multiple options are provided for prediction including on response and parameter scale, by quantiles, cdf, density or probability, and reporting by marginal effect.

```{r applied-answers}
# Predict the response on response scale with confidence interval
predict(
  fit,
  newdata = dat,
  type = "response",
  se.fit = TRUE,
  interval = "confidence"
)

# Predict the estimated parameter values for mu, sigma etc.
predict(
  fit,
  newdata = dat,
  type = "parameters"
)

# Predict specified quantiles
predict(
  fit,
  newdata = dat,
  type = "quantile",
  probs = c(0.1, 0.5, 0.9)
)

# Predict cdf
predict(
  fit,
  newdata = dat,
  type = "cdf",
  q = 10
)

# Predict probability density
predict(
  fit,
  newdata = dat,
  type = "density",
  y = 10
)

# Predict observation probability (either above or below response value)
predict(
  fit,
  newdata = dat,
  type = "probability",
  q = 10,
  direction = "above"
)

# Calculate average response across a covariate for the margin
marginal_effects(
  fit,
  newdata = dat,
  variable = "treatment",
  parameter = "mu",
  se.fit = TRUE
)
```

Simulate from the fitted model as desired, for a total of `nsim` replications. Note that simulation uses both copula and margin structure so dependence is maintained in simulations.

```{r simulation}
sim_model <- simulate(fit, nsim = 100, seed = 1)
```

## Optional: Compare/benchmark to familiar models: gee (geepack), glmm (lme4), gam (mgcv)

Since we know most users would be considering a longitudinal gamlss model as an alternative to existing solutions using gees, gam or glmms we provide easy to use comparison tools to benchmark longitudinal gamlss models to existing methods.

```{r comparator}
# Fit comparison models and summarise fits for a model with linear age term
bench_nosmooth <- benchmark_standard_models(
  data = dat,
  formula = response ~ treatment + time + age_scaled,
  subject_var = "subject",
  family = "gamma",
  fit = fit_reduced
)

print(bench_nosmooth)

# Fit comparison models and summarise fits for a model with smooth age term
bench_smooth <- benchmark_standard_models(
  data = dat,
  formula = response ~ treatment + time + s(age_scaled),
  subject_var = "subject",
  family = "gamma",
  fit = fit
)

print(bench_smooth)

```

In general, note that the purpose of the gamlss.longitudinal model is to accurately capture the full joint distribution shape. If the item of interest is just the mean, then gamlss.longitudinal is unlikely to provide large gains except for where distributions are clearly non-standard. By modelling the whole joint outcome distribution we can more accurately answer quesitons such as whether quantiles, tail risks, or uncertainty (SEs, confidence intervals, prediction intervals) change in a meaningful way. In addition, note that the consumption of degrees of freedom for calculation of accurate conditional AICs for random effect models can be substantial compared to the small number of parameters needed to fit a copula-based dependence structure.

We provide additional information on comparisons in other vignettes: `vignette("benchmarking-adoption", package = "gamlss.longitudinal")` for scenario-level benchmark reporting.

##

For a more detailed workthrough example, including interpretation of model fit and diagnostics, please review the [Detailed worked example](native-simulation-workflow.html).
