---
title: "Standard Workflow API, Inference, and Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard Workflow API, Inference, and Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette is the recommended starting workflow for applied longitudinal
analyses with `gamlss.longitudinal`. It is deliberately narrow: get the margin
roughly right, choose a plausible dependence model, fit the longitudinal model,
check the result, then report response-scale answers.

Use this as the default path before moving to the focused guides for replacing
a GEE, replacing a random-intercept GLMM, handling non-Gaussian outcomes,
missing visits, or time-varying dependence.

## Functions Covered

This vignette demonstrates or describes every user-facing function added for
the standard adoption workflow.

| Function | Where it appears | Purpose |
|---|---|---|
| `screen_margin()` | Step 1 | Screen candidate marginal distributions before fitting dependence. |
| `print.margin_screen()` | Step 1 | Automatically prints the ordered margin screen. |
| `fit_longitudinal()` | Step 3 | Fit the recommended GAMLSS-copula workflow with adoption-friendly defaults. |
| `confint.gamlss.longitudinal()` | Step 3 | Report coefficient confidence intervals through the standard `confint()` generic. |
| `wald_test()` | Step 3 | Run individual, joint, or contrast-matrix Wald tests. |
| `likelihood_compare()` | Step 3 | Compare nested or approximately nested longitudinal fits. |
| `bootstrap_inference()` | Step 3 | Run parametric simulation-and-refit coefficient uncertainty checks. |
| `print.gamlss_longitudinal_wald_test()` | Step 3 | Automatically prints `wald_test()` output. |
| `print.gamlss_longitudinal_likelihood_compare()` | Step 3 | Automatically prints likelihood comparison output. |
| `print.gamlss_longitudinal_bootstrap()` | Step 3 | Automatically prints bootstrap summaries. |
| `check_model()` | Step 4 | Return decision-oriented diagnostics, warnings, and a model-role recommendation. |
| `print.gamlss_longitudinal_check()` | Step 4 | Automatically prints the diagnostic object. |
| `plot.gamlss_longitudinal_check()` | Step 4 | Draw stored PIT, QQ, worm, and rootogram diagnostics through `plot(check)`. |
| `predict.gamlss.longitudinal()` | Step 5 | Return means, parameters, quantiles, CDFs, densities, probabilities, and uncertainty intervals. |
| `simulate.gamlss.longitudinal()` | Step 5 | Simulate marginal or copula-preserving fitted responses. |
| `marginal_effects()` | Step 5 | Summarize counterfactual response-scale effects. |
| `effects.gamlss.longitudinal()` | Step 5 | Provide a familiar `effects()` alias for `marginal_effects()`. |

## Data Requirements

Start with one row per subject and visit. The model needs:

- a response column;
- a subject identifier;
- a visit, time, or occasion variable;
- scientific predictors for the margin;
- optional predictors for scale, shape, or dependence.

```{r data-shape}
str(dat)
table(dat$subject)
table(dat$time)
```

Unbalanced panels and missing visits are allowed, but the missingness mechanism
still needs to be part of the scientific judgement. See
`vignette("missing-visits", package = "gamlss.longitudinal")` for that case.

## Step 1: Screen The Margin

Screen the response distribution before fitting the longitudinal dependence
model. This is not the final model; it is a fast way to avoid starting with an
obviously poor family.

```{r margin-screen}
margin_screen <- screen_margin(
  dat,
  response_var = "response",
  families = c("NO", "GA", "LOGNO", "PO")
)

margin_screen
```

Use the screen as a shortlist, then choose a family that makes scientific sense.
For example, positive skewed outcomes often start with `GA()` or `LOGNO()`,
counts with `PO()`, and approximately symmetric continuous outcomes with
`NO()`.

```{r margin-family}
margin_dist <- gamlss.dist::GA(mu.link = "log", sigma.link = "log")
```

## Step 2: Screen The Copula

Use `select_copula()` to compare dependence families. If pseudo-observations
are already available, pass them through `u_var`. In a real analysis, they
usually come from a preliminary marginal model or a workflow-specific
pre-processing step.

```{r copula-screen}
copula_screen <- select_copula(
  data = dat,
  u_var = "u",
  subject_var = "subject",
  time_var = "time",
  families = c("N", "C", "F", "G", "J", "t")
)

copula_screen
copula_dist <- attr(copula_screen, "selected")
```

The selected family is a starting point, not a guarantee. Dependence diagnostics
after the full fit matter more than the screen.

## Step 3: Fit The Longitudinal Model

Put the main scientific effects in `mu.formula`. Add `sigma.formula` when spread
changes across time, treatment, or baseline covariates. Add `theta.formula` when
dependence plausibly changes over follow-up or with subject-level predictors.

```{r fit}
fit <- fit_longitudinal(
  dataset = dat,
  margin_dist = margin_dist,
  copula_dist = copula_dist,
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + s(age_scaled, bs = "ps"),
  sigma.formula = ~ treatment + time,
  theta.formula = ~ time,
  compute_vcov = FALSE,
  run_checks = TRUE
)
```

For final reporting, request variance-covariance output on the final model:

```{r final-vcov}
summary(fit, include_vcov = TRUE)
confint(fit)
wald_test(fit, terms = "mu.treatment")
likelihood_compare(fit_reduced, fit)

boot <- bootstrap_inference(fit, R = 200, terms = "mu.treatment")
boot
```

See `vignette("inference-uncertainty", package = "gamlss.longitudinal")` for
the reporting language around Hessian, numerical-Hessian, delta-method, and
simulation-based uncertainty.
See `vignette("diagnostics-decisions", package = "gamlss.longitudinal")` for
the decision language around marginal, tail, dependence, and inference warnings.

## Step 4: Check Before Interpreting

Run `check_model()` before interpreting coefficients or predictions.

```{r check}
check <- check_model(fit, include_vcov = TRUE)
check
check$warnings
check$recommendation
plot(check)
copula_time_summary(fit)
```

Treat the output as a decision aid:

- marginal warnings mean the response family or parameter formulas need work;
- tail warnings mean quantile or probability claims are not yet trustworthy;
- residual-dependence warnings mean the copula family or `theta.formula` may be
  underfitting longitudinal association;
- inference warnings mean intervals should be reported as approximate or the
  model should be simplified.

Only move to primary-model reporting once the diagnostics match the intended
estimand.

## Step 5: Report Applied Answers

Prefer response-scale summaries over raw coefficient tables when communicating
with applied collaborators.

```{r applied-answers}
predict(
  fit,
  newdata = dat,
  type = "response",
  se.fit = TRUE,
  interval = "confidence"
)

predict(
  fit,
  newdata = dat,
  type = "parameters"
)

predict(
  fit,
  newdata = dat,
  type = "quantile",
  probs = c(0.1, 0.5, 0.9)
)

predict(
  fit,
  newdata = dat,
  type = "cdf",
  q = 10
)

predict(
  fit,
  newdata = dat,
  type = "density",
  y = 10
)

predict(
  fit,
  newdata = dat,
  type = "probability",
  q = 10,
  direction = "above"
)

marginal_effects(
  fit,
  newdata = dat,
  variable = "treatment",
  parameter = "mu",
  se.fit = TRUE
)

effects(
  fit,
  newdata = dat,
  variable = "treatment",
  parameter = "mu",
  se.fit = TRUE
)
```

If the target is a trajectory or a dependence-sensitive event, simulate from the
fitted model:

```{r simulation}
sim_marginal <- simulate(fit, nsim = 100, type = "marginal", seed = 1)
sim_copula <- simulate(fit, nsim = 100, type = "copula", seed = 1)
```

## Step 6: Compare Against Familiar Models

A replacement claim needs a sensitivity comparison. Use the built-in benchmark
scaffold for a quick check, then the adoption benchmark runner for repeated
simulation evidence.

```{r comparator}
benchmark_standard_models(
  data = dat,
  formula = response ~ treatment + time + age_scaled,
  subject_var = "subject",
  family = "gaussian",
  comparators = c("gee", "glmm", "gam")
)
```

For repeated evidence:

```{r adoption-benchmark}
bench <- run_adoption_benchmarks(
  scenarios = "gaussian_heteroskedastic",
  reps = 25,
  methods = c("rs_separate", "gee", "glmm", "gam"),
  write_results = TRUE
)

write_benchmark_report(bench)
```

Use `vignette("benchmarking-adoption", package = "gamlss.longitudinal")` for
scenario-level benchmark reporting.

## Minimal Reporting Checklist

A complete applied report should state:

1. why a distributional longitudinal model was needed;
2. the margin family and copula family;
3. which predictors entered `mu`, `sigma`, and `theta`;
4. the main `check_model()` decisions;
5. the inference route used for intervals;
6. the response-scale estimands reported;
7. the GEE, GLMM, or GAM sensitivity result.

If the report only needs a marginal mean and the diagnostics do not suggest
distributional or dependence problems, a GEE may still be the better primary
model. The goal is not to replace familiar models everywhere; it is to make the
GAMLSS-copula workflow the standard when margins, tails, heteroskedasticity, or
dependence are part of the answer.
