---
title: "Time-Varying Dependence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-Varying Dependence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide is for longitudinal settings where repeated measurements are not
equally dependent across follow-up. A random intercept or exchangeable working
correlation treats within-subject association as essentially constant. A
GAMLSS-copula model can instead put time or other predictors in the copula
dependence formula.

## When To Consider It

Time-varying dependence is worth checking when:

- measurements close to baseline are more strongly associated than later
  measurements,
- treatment or disease stage changes within-subject clustering,
- residual dependence remains after a constant-dependence model,
- tail clustering changes over follow-up,
- the scientific question involves persistence, recovery, or episodic risk.

Use a constant `theta.formula = ~ 1` as the first dependence model. Move to
`theta.formula = ~ time` only when the data have enough adjacent pairs to
support the extra parameter.

## Fit A Constant-Dependence Model First

```{r constant-fit}
fit_constant <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = gamlss.dist::NO(),
  copula_dist = "N",
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + age_scaled,
  sigma.formula = ~ treatment + time,
  theta.formula = ~ 1
)

check_constant <- check_model(fit_constant)
check_constant$residual_dependence
```

If residual dependence remains, fit a time-varying dependence model:

```{r time-fit}
fit_time <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = gamlss.dist::NO(),
  copula_dist = "N",
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + age_scaled,
  sigma.formula = ~ treatment + time,
  theta.formula = ~ time
)

check_time <- check_model(fit_time)
check_time$residual_dependence
check_time$copula
copula_time_summary(fit_time)
```

For more complex patterns, use smooth terms cautiously:

```{r smooth-time-fit}
fit_smooth_time <- gamlss_longitudinal(
  dataset = dat,
  margin_dist = gamlss.dist::NO(),
  copula_dist = "N",
  time_var = "time",
  subject_var = "subject",
  mu.formula = response ~ treatment + time + age_scaled,
  sigma.formula = ~ treatment + time,
  theta.formula = ~ s(time, bs = "ps", k = 5)
)
```

Smooth dependence formulas need enough adjacent pairs across the time range.
If the Hessian, convergence, or diagnostics become unstable, prefer a simpler
linear or constant dependence model.

## Interpret Dependence Separately From The Mean

Changing `theta` does not mean the response mean is changing. It means the
within-subject association between adjacent observations changes after
accounting for the marginal distribution. Report it as a dependence result:

```{r interpretation}
summary(fit_time)
copula_time_summary(fit_time)
plot_copula_diagnostics(fit_time, data = dat)
```

Good applied wording is narrow:

- "The fitted dependence is stronger earlier in follow-up."
- "After modelling the marginal mean and scale, residual dependence is reduced
  by allowing the copula parameter to vary with time."
- "The time-varying dependence term is exploratory because adjacent pairs are
  sparse after visit 4."

## Benchmark The Claim

The adoption benchmark suite includes a time-varying dependence scenario:

```{r benchmark}
dep_bench <- run_adoption_benchmarks(
  scenarios = "time_varying_dependence",
  reps = 50,
  methods = c("rs_separate", "gee", "glmm", "gam"),
  max_elapsed_sec = 60,
  write_results = TRUE
)

dep_bench$summary$summary
```

The key scenario metric is `benchmark_theta_time_abs_error`, which compares the
estimated time effect in the copula parameter with the simulated truth on the
link scale. Standard GEE/GLMM/GAM comparator rows do not estimate this quantity,
so they are useful for runtime and mean prediction context rather than direct
dependence recovery.

## When To Keep Dependence Constant

Keep `theta.formula = ~ 1` when there are few visits, many missing adjacent
pairs, weak residual dependence, or unstable convergence. A constant-dependence
model with good diagnostics is usually easier to explain and defend than a
time-varying dependence model that is only weakly identified.
