---
title: "Joint versus Separate Optimization: Results Review"
subtitle: "Module 03 candidate selection and expanded review simulation"
date: "2026-06-12"
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    geometry:
      - margin=18mm
execute:
  echo: false
  warning: false
  message: false
---

```{r}
candidate_path <- file.path(
  "..", "..", "results", "jss-replication", "expanded", "data",
  "03-joint-vs-separate-optimization-candidate-selection.csv"
)
results_path <- file.path(
  "..", "..", "results", "jss-replication", "expanded", "data",
  "03-joint-vs-separate-optimization-results.csv"
)
summary_path <- file.path(
  "..", "..", "results", "jss-replication", "expanded", "tables",
  "03-joint-vs-separate-optimization-summary.csv"
)
metric_path <- file.path(
  "..", "..", "results", "jss-replication", "expanded", "tables",
  "03-joint-vs-separate-optimization-metric-wins.csv"
)
delta_figure <- file.path(
  "..", "..", "results", "jss-replication", "expanded", "figures",
  "03-joint-vs-separate-optimization-deltas.png"
)
dashboard_figure <- file.path(
  "..", "..", "results", "jss-replication", "expanded", "figures",
  "03-joint-vs-separate-optimization-metric-dashboard.png"
)

candidates <- read.csv(candidate_path, stringsAsFactors = FALSE)
results <- read.csv(results_path, stringsAsFactors = FALSE)
metric_wins <- read.csv(metric_path, stringsAsFactors = FALSE)

fmt <- function(x, digits = 3) {
  ifelse(is.na(x), "", format(signif(x, digits), trim = TRUE, scientific = FALSE))
}

metric_label <- c(
  benchmark_mean_rmse = "Mean RMSE",
  benchmark_mean_mae = "Mean MAE",
  benchmark_mean_bias = "Mean bias",
  benchmark_rmse = "Observed RMSE",
  benchmark_mae = "Observed MAE",
  benchmark_q90_mae = "90th percentile MAE",
  benchmark_neg_log_score = "Negative log score",
  benchmark_upper_tail_error_90 = "90th percentile upper-tail error",
  benchmark_theta_time_abs_error = "Theta time absolute error",
  benchmark_interval_coverage_95 = "95% interval coverage",
  benchmark_interval_width_95 = "95% interval width",
  benchmark_pit_mean_abs_error = "PIT mean absolute error",
  benchmark_pit_ks_p_value = "PIT KS p-value",
  benchmark_tail_error_lower_05 = "Lower 5% tail error",
  benchmark_tail_error_upper_05 = "Upper 5% tail error",
  elapsed_sec = "Runtime"
)

score_rule <- c(
  benchmark_interval_coverage_95 = "closer to 0.95",
  benchmark_pit_ks_p_value = "higher",
  elapsed_sec = "lower"
)

pair_key_cols <- c(
  "joint_review_rep", "family", "copula", "design", "n_subject", "n_time",
  "dependence", "missingness", "start_mode"
)
pair_key <- interaction(results[pair_key_cols], drop = TRUE, lex.order = TRUE)
deltas <- do.call(rbind, lapply(unique(pair_key), function(key) {
  x <- results[pair_key == key, , drop = FALSE]
  joint <- x[x$method == "rs_joint", , drop = FALSE][1, , drop = FALSE]
  separate <- x[x$method == "rs_separate", , drop = FALSE][1, , drop = FALSE]
  data.frame(
    family = joint$family,
    copula = joint$copula,
    design = joint$design,
    delta_joint_loglik = joint$joint_loglik - separate$joint_loglik,
    delta_elapsed_sec = joint$elapsed_sec - separate$elapsed_sec,
    delta_mean_rmse = joint$benchmark_mean_rmse - separate$benchmark_mean_rmse,
    delta_neg_log_score = joint$benchmark_neg_log_score - separate$benchmark_neg_log_score,
    delta_theta_time_abs_error = joint$benchmark_theta_time_abs_error -
      separate$benchmark_theta_time_abs_error,
    stringsAsFactors = FALSE
  )
}))

run_status <- aggregate(
  cbind(success = success == TRUE, converged = converged == TRUE, elapsed_sec = elapsed_sec) ~ method,
  data = results,
  FUN = function(x) {
    if (is.logical(x)) sum(x) else stats::median(x, na.rm = TRUE)
  }
)
run_status$n <- as.integer(table(results$method)[run_status$method])
run_status <- run_status[, c("method", "n", "success", "converged", "elapsed_sec")]
names(run_status) <- c("Method", "Fits", "Successful", "Marked converged", "Median seconds")

delta_summary <- data.frame(
  Metric = c(
    "Joint log-likelihood",
    "Negative log score",
    "Mean RMSE",
    "Theta time absolute error",
    "Runtime"
  ),
  Direction = c("higher is better", "lower is better", "lower is better", "lower is better", "lower is better"),
  `Finite pairs` = c(
    sum(is.finite(deltas$delta_joint_loglik)),
    sum(is.finite(deltas$delta_neg_log_score)),
    sum(is.finite(deltas$delta_mean_rmse)),
    sum(is.finite(deltas$delta_theta_time_abs_error)),
    sum(is.finite(deltas$delta_elapsed_sec))
  ),
  `Joint better` = c(
    sum(deltas$delta_joint_loglik > 0, na.rm = TRUE),
    sum(deltas$delta_neg_log_score < 0, na.rm = TRUE),
    sum(deltas$delta_mean_rmse < 0, na.rm = TRUE),
    sum(deltas$delta_theta_time_abs_error < 0, na.rm = TRUE),
    sum(deltas$delta_elapsed_sec < 0, na.rm = TRUE)
  ),
  `Mean delta` = c(
    mean(deltas$delta_joint_loglik, na.rm = TRUE),
    mean(deltas$delta_neg_log_score, na.rm = TRUE),
    mean(deltas$delta_mean_rmse, na.rm = TRUE),
    mean(deltas$delta_theta_time_abs_error, na.rm = TRUE),
    mean(deltas$delta_elapsed_sec, na.rm = TRUE)
  ),
  `Median delta` = c(
    stats::median(deltas$delta_joint_loglik, na.rm = TRUE),
    stats::median(deltas$delta_neg_log_score, na.rm = TRUE),
    stats::median(deltas$delta_mean_rmse, na.rm = TRUE),
    stats::median(deltas$delta_theta_time_abs_error, na.rm = TRUE),
    stats::median(deltas$delta_elapsed_sec, na.rm = TRUE)
  ),
  check.names = FALSE
)

candidate_display <- candidates[, c(
  "candidate_id", "review_role", "copula", "scenario", "profile_id",
  "source_name", "delta_train_joint_loglik_mean",
  "delta_test_log_score_per_obs_mean", "delta_train_rmse_tau_mean"
)]
names(candidate_display) <- c(
  "ID", "Role", "Copula", "Scenario", "Profile", "Source",
  "Train joint LL delta", "Held-out log score delta/obs", "Tau RMSE delta"
)
candidate_display$Scenario[is.na(candidate_display$Scenario)] <- ""
candidate_display$Profile[is.na(candidate_display$Profile)] <- ""
candidate_display$`Train joint LL delta` <- fmt(candidate_display$`Train joint LL delta`)
candidate_display$`Held-out log score delta/obs` <- fmt(candidate_display$`Held-out log score delta/obs`)
candidate_display$`Tau RMSE delta` <- fmt(candidate_display$`Tau RMSE delta`)

metric_pivot <- reshape(
  metric_wins[, c("metric", "estimand", "method", "n_finite", "wins", "ties", "losses", "win_or_tie_rate")],
  idvar = c("metric", "estimand"),
  timevar = "method",
  direction = "wide"
)
metric_pivot$Metric <- ifelse(
  metric_pivot$metric %in% names(metric_label),
  metric_label[metric_pivot$metric],
  metric_pivot$metric
)
metric_pivot$`Joint win/tie` <- fmt(as.numeric(metric_pivot$win_or_tie_rate.rs_joint) * 100)
metric_pivot$`Separate win/tie` <- fmt(as.numeric(metric_pivot$win_or_tie_rate.rs_separate) * 100)
metric_pivot$`Joint W-T-L` <- paste(metric_pivot$wins.rs_joint, metric_pivot$ties.rs_joint, metric_pivot$losses.rs_joint, sep = "-")
metric_pivot$`Separate W-T-L` <- paste(metric_pivot$wins.rs_separate, metric_pivot$ties.rs_separate, metric_pivot$losses.rs_separate, sep = "-")
metric_display <- metric_pivot[, c("Metric", "estimand", "Joint win/tie", "Separate win/tie", "Joint W-T-L", "Separate W-T-L")]
names(metric_display)[2] <- "Estimand"
```

# Executive Summary

This report reviews the current `03-joint-vs-separate-optimization` module.
The module now has two layers of evidence:

1. A historical candidate-selection pass that mines retained exploratory summaries.
2. A fresh expanded review simulation comparing `rs_joint` and `rs_separate`.

The result is not a simple "`joint` always wins" story. The current evidence is
more useful and more defensible by estimand:

- Joint optimization is strongest for dependence-sensitive and distributional
  targets, including theta-time recovery, PIT calibration, quantiles, and lower
  tail behavior.
- Separate optimization is faster and has cleaner convergence status in this
  run.
- Direct mean/observed-response metrics are mixed, with neither method
  dominating across all cases.

# Run Configuration and Outputs

The expanded review run used:

- families: `NO`, `NBI`;
- copulas: `N`, `C`, `G`;
- designs: `intercept`, `covariate`, `time_dependence`;
- subjects: `48`;
- repeated time points: `1:4`;
- methods: `rs_separate`, `rs_joint`;
- replicate count: `10` per family/copula/design cell;
- maximum outer iterations: `8`;
- maximum inner iterations: `8`;
- per-fit elapsed-time limit: `45` seconds.

The generated outputs are:

- `results/jss-replication/expanded/data/03-joint-vs-separate-optimization-candidate-selection.csv`
- `results/jss-replication/expanded/data/03-joint-vs-separate-optimization-results.csv`
- `results/jss-replication/expanded/tables/03-joint-vs-separate-optimization-summary.csv`
- `results/jss-replication/expanded/tables/03-joint-vs-separate-optimization-metric-wins.csv`
- `results/jss-replication/expanded/figures/03-joint-vs-separate-optimization-deltas.png`
- `results/jss-replication/expanded/figures/03-joint-vs-separate-optimization-metric-dashboard.png`

# Candidate Selection

The historical candidate-mining step selected seven review cases:

- five `joint_win` cases;
- one `tie_control` case;
- one `cautionary` case.

```{r}
knitr::kable(candidate_display, booktabs = TRUE, longtable = TRUE)
```

The five `joint_win` cases were selected because historical summaries showed
positive gains in held-out log score and training joint log-likelihood, usually
with reductions in dependence RMSE. The cautionary case is useful because it
has likelihood/log-score gains but worse tau RMSE, which prevents the paper
from overclaiming that likelihood improvement is automatically better for every
estimand.

# Fresh Expanded Simulation

The fresh expanded run produced 360 model fits: 180 paired cases with one
`rs_separate` and one `rs_joint` fit in each case.

```{r}
run_status_print <- run_status
run_status_print$`Median seconds` <- fmt(run_status_print$`Median seconds`)
knitr::kable(run_status_print, booktabs = TRUE)
```

Separate optimization completed successfully in all 180 fits and was marked
converged in all 180 fits. Joint optimization completed successfully in 174 of
180 fits and was marked converged in 103 of 180 fits. The non-successful joint
fits were retained as missing metric rows in the win/tie/loss summaries.

Runtime favored separate optimization. Median runtime was about 0.26 seconds
for separate optimization in the 1-replicate run; in the 10-replicate run the
median was about 0.62 seconds for separate optimization and about 1.57 seconds
for joint optimization.

# Joint-minus-Separate Deltas

Positive deltas mean joint is larger than separate. For error and runtime
metrics, negative deltas are better for joint.

```{r}
delta_summary_print <- delta_summary
delta_summary_print$`Mean delta` <- fmt(delta_summary_print$`Mean delta`)
delta_summary_print$`Median delta` <- fmt(delta_summary_print$`Median delta`)
knitr::kable(delta_summary_print, booktabs = TRUE)
```

The 10-replicate run shows:

- joint optimization retained a clear advantage for theta-time dependence
  recovery;
- negative log score and calibration became slightly less favorable to joint
  than in the 1-replicate check;
- mean RMSE remained essentially mixed;
- theta-time absolute error was better for joint in 35 of 60 finite
  time-dependence cases;
- runtime was better for joint in only 10 of 180 paired cases.

# Metric Win/Tie Summary

Win/tie rates below are percentages. They are computed case-by-case using the
benchmark scoring rules.

```{r}
knitr::kable(metric_display, booktabs = TRUE, longtable = TRUE)
```

Important patterns:

- Dependence: joint had the stronger result for `benchmark_theta_time_abs_error`
  with a 66.7 percent win/tie rate versus 33.3 percent for separate.
- Calibration: joint was slightly stronger on PIT metrics.
- Quantiles and tails: joint was stronger on `benchmark_q90_mae` and lower-tail
  error; upper-tail behavior was close.
- Mean and observed-response fit: mixed. Separate was stronger on observed MAE
  and observed RMSE win/tie rates; joint was slightly stronger on mean MAE.
- Runtime: separate clearly dominated.

# Figures

```{r, out.width="100%"}
knitr::include_graphics(delta_figure)
```

```{r, out.width="100%"}
knitr::include_graphics(dashboard_figure)
```

# Interpretation for the Paper

The most defensible interpretation is by estimand rather than by a single
overall winner.

Joint optimization is worth emphasizing when the paper discusses dependence
recovery and distributional fit. It is especially relevant for settings where
copula/dependence estimation affects calibration, quantiles, and tail behavior.

Separate optimization remains a strong baseline. It is faster, has cleaner
convergence status in this run, and remains competitive on mean and
observed-response metrics.

The paper should therefore avoid saying that joint optimization universally
dominates. A stronger claim is:

> Joint optimization can improve dependence-sensitive and distributional
> estimands, but those gains come with runtime and convergence-status costs.
> Comparisons should be reported by estimand.

# Limitations and Next Checks

This run is a review-scale expanded run with 10 replicates per
family/copula/design cell. The module can be rerun with more replicates by
setting `GAMLSS_LONGITUDINAL_JSS_JVS_REPS`, for example to `100`.

Recommended next checks:

- rerun the expanded module with more replicates for the 7 selected candidate
  scenarios;
- inspect why successful joint fits are less often marked converged;
- decide whether the cautionary case should be included in the manuscript or
  kept for supplementary discussion;
- add a focused table that separates dependence/calibration metrics from mean
  and runtime metrics.
